```{css echo=F}
pre {
  overflow-x: scroll;
}
pre code {
  white-space: pre;
}
```

# Setup
```{r setup, message=F}
library(data.table)
library(readxl)

options(datatable.na.strings=c('NA',''))

if(!file.exists('phenotypes/')) {
  system('gsutil -m cp -r gs://fc-secure-4a392455-5587-4d6f-b8bd-01a1f834ae63/phenotypes .')
  system('gsutil    cp    gs://fc-secure-4b3e979d-ba8b-43c4-a5ec-b41ab42ce606/MesaMetabolomics_PilotX01_2023_02_16/MESA_PhenoUsmanSIDNO_20230214.txt phenotypes')
}

winsorizeBySd <- function(x, SDs=5) {
  lower_bound <- mean(x,na.rm=T) - SDs*sd(x,na.rm=T)
  upper_bound <- mean(x,na.rm=T) + SDs*sd(x,na.rm=T)
  print(paste(sum(x<lower_bound,na.rm=T), 'values winsorized at the lower bound.'))
  print(paste(sum(x>upper_bound,na.rm=T), 'values winsorized at the upper bound.'))
  x[x<lower_bound] <- lower_bound
  x[x>upper_bound] <- upper_bound
  return(x)
}
```
```{r merge-fxns, echo=F}
# Merge which coalesces columns with duplicated names along the way.
# Note: when coalescing same-name cols of x and y, x's values will be chosen first.
mergecoalesce <- \(x,y, by=NULL, by.x=NULL, by.y=NULL, ..., force_coalesce=F) {
  if(!is.null(by)) by.x <- by.y <- by
  setkeyv(x,by)
  setkeyv(y,by)
  merged <- merge(x,y, by.x=by.x, by.y=by.y, ...)

  nm_dups <- intersect(names(x),names(y)) |> setdiff(c(by.x,by.y))

  for(nm in nm_dups) {
    nms_w_suffix <- paste0(nm, c('.x','.y'))

    classes <- merged[, sapply(.SD,class), .SDcols=nms_w_suffix]
    if(length(unique(classes))!=1L) { # Danger, cols' classes don't match
      if(!force_coalesce) {
        warning('Not coalescing these columns because their class doesn\'t match:\n', paste(capture.output(print(classes)), collapse='\n'))
        next
      } else { # Force coalescing anyway (cast to character for maximum compatibility)
        merged[, (nms_w_suffix) := lapply(.SD, as.character), .SDcols = nms_w_suffix]
    }}

    merged <- merged[
      ][, (nm) := fcoalesce(.SD), .SDcols= nms_w_suffix 
      ][, .SD                   , .SDcols=!nms_w_suffix]
  }

  merged
}

coalesce_samenames <- \(dt, force_coalesce=F) {
  nms <- unique(colnames(dt)[duplicated(colnames(dt))])
  ns  <- sapply(nms,\(nm) sum(nm==colnames(dt)))
  exact_nms <- paste0('^',nms,'$')

  for(i in seq_along(nms)) {
    nm <- nms[i]
    n  <- ns [i]
    exact_nm <- exact_nms[i]

    classes <- dt[, sapply(.SD,class), .SDcols=patterns(exact_nm)]
    if(length(unique(classes))!=1L) { # Danger, cols' classes don't match
      if(!force_coalesce) { 
        warning('Not coalescing these columns because their class doesn\'t match:\n', paste(capture.output(print(classes)), collapse='\n'))
        next
      } else { # Force coalescing anyway (cast to character for maximum compatibility)
        cols_as.char <- dt[, lapply(.SD,as.character), .SDcols=patterns(exact_nm)]
        coalesced_col <- fcoalesce(cols_as.char)
      }
    } else { # All is well, cols' classes match
      coalesced_col <- dt[, fcoalesce(.SD), .SDcols=patterns(exact_nm)]
    }

    for(j in 1:n) dt[, (nm) := NULL]
    dt[, (nm) := coalesced_col]
  }
  dt
}
```

# Join all datasets
```{r join, cache=T}
# Create one dt containing all the IDs we would ever want to annotate.
# That is, TOM_Ids with metabolomics info, and mesa_ids with PA info. Note that one mesa_id may correspond to multiple TOM_Ids.
# Then, can merge with impunity using all.x=T for everything, without fear of losing any info.
data <-
  merge(fread('phenotypes/primary_phenotypes.csv')           [, mesa_id := sub('phs000209.v13_', '', `\\_Parent Study Accession with Subject ID\\`)],
        fread('phenotypes/MESA_PhenoUsmanSIDNO_20230214.txt')[, mesa_id := as.character(sidno)][, NWD_Id := fcoalesce(nwdid2,nwdid1)],
        by='mesa_id', all=T) |>
  melt(measure.vars=c('tomid1','tomid2','tomid3','tomid4','tomid5'),
       variable.name='metabolomics_exam',
       value.name='TOM_Id')
data <- data[
  ][!(duplicated(mesa_id) & is.na(TOM_Id)) # Melt also expanded rows where tomid#==NA. Rm these mesa_ids associated with multiple NA TOM_Ids.
  ][is.na(TOM_Id), metabolomics_exam := NA
  ][,              metabolomics_exam := sub('^tomid','',metabolomics_exam)
]

data <- 
Reduce(\(x,l) mergecoalesce(x, l$y, by=l$by, force_coalesce=T, all.x=T),
  list(                x=data,
    list(by='mesa_id', y=fread('phenotypes/covariates.csv')                             [, mesa_id := sub('phs000209.v13_', '', `\\_Parent Study Accession with Subject ID\\`)]),
    list(by='mesa_id', y=fread('phenotypes/nmr_metabolites.csv')                        [, mesa_id := sub('phs000209.v13_', '', `\\_Parent Study Accession with Subject ID\\`)]),
    list(by='mesa_id', y=fread('phenotypes/basic_phenotypes.csv')                       [, mesa_id := sub('phs000209.v13_', '', `\\_Parent Study Accession with Subject ID\\`)]),
    list(by='mesa_id', y=fread('phenotypes/mesa5_phenos_diet.csv')                      [, mesa_id := sub('phs000209.v13_', '', `\\_Parent Study Accession with Subject ID\\`)]),
    list(by='mesa_id', y=fread('phenotypes/mesa5_phenos_basic.csv')                     [, mesa_id := sub('phs000209.v13_', '', `\\_Parent Study Accession with Subject ID\\`)]),
    list(by='mesa_id', y=fread('phenotypes/id_match_file.csv')                          [, mesa_id := as.character(Cohort_Specific_Id)][!duplicated(mesa_id)]                  ), # Duplicated mesa_ids due to TOR replicates. We aren't using RNAseq data so just pick whichever.
    list(by='mesa_id', y=fread('phenotypes/freeze9b_sample_annot_2020-08-20.txt')       [, mesa_id := subject_id][!duplicated(mesa_id)]                                        ),
    list(by='mesa_id', y=fread('phenotypes/SHARe_AncilMesaNMR_LP4_DS.txt')              [, mesa_id := as.character(subject_id)]                                                ),
    list(by='mesa_id', y=setDT(read_excel('phenotypes/draw_dates_MESA_study_site.xlsx'))[, mesa_id := sidno]                                                                   ),
    list(by= 'NWD_Id', y=fread('genotypes.csv')                                         [, NWD_Id := NWD_ID]                                                                   ),
    list(by= 'NWD_Id', y=fread('phenotypes/freeze9_pcair_results.tsv')                  [, NWD_Id := sample.id]                                                                ),
    list(by= 'TOM_Id', y=fread('metabolomics/sample_info.csv')                          [, TOM_Id := sample_id]                                                                ),
    list(by= 'TOM_Id', y=fread('metabolomics/QCd/merged_QCd.csv')                                                                                                              )) # Merge metabolomics last b/c it adds tons of cols
)
```

# Rename columns
```{r rename}
data <- data |>
setnames(\(nm)  sub('.*\\\\(.*)\\\\$', '\\1', nm)) |> # Fixes some names e.g. '\\phs000209\\pht001116\\phv00084442\\age1c\\' -> 'age1c'
setnames(\(nm) gsub(      '\\\\'     ,  '',   nm)) |> # Rm remaining backslashes
setnames(\(nm)  sub('^PC'            , 'gPC', nm)) |> # 'PC#' -> 'gPC#' To distinguish genetic PCs from metabolite PCs we'll add later
coalesce_samenames(force_coalesce=T)                  # Merge any columns whose names are now duplicated after renaming stuff
data <- data[, `:=`(                                  # Manually pretty up some names
   # SHARe...txt      #primary...csv    #nmr...csv           #covariates.csv           
   HDL_C=nhdlc1,      hdl=hdl1,         HDL_C_lp3=nhdlc31c,  ses_score=F1_PC2_1,
   HDL_P=chdlp1,      ldl=ldl1,         HDL_P_lp3= hdlp31c,  drinks_per_week=alcwkc1,
   HDL_size=hdlz1,    chol=chol1,       HDL_size_lp3=hz31,   ahei_score=ahei_2010_1,
   L_HDL_P=l_chdlp1,  tg=trig1,         L_HDL_P_lp3=hl31,    dash_score=dash_sodium1,
   M_HDL_P=m_chdlp1,      pa=exercm1c,  M_HDL_P_lp3=hm31,
   S_HDL_P=s_chdlp1,  mod_pa= pamcm1c,  S_HDL_P_lp3=hs31,    #basic...csv
   H1P=h1p1,          vig_pa= pavcm1c,                       age=age1c,
   H2P=h2p1,            mvpa=pamvcm1c,                       bmi=bmi1c,
   H3P=h3p1,  #id_match_file.csv         #draw_dates...xlsx
   H4P=h4p1,  prop_African=African,      month=`exam1:month`,   
   H5P=h5p1,  prop_American=American,    season=`exam1:season`, 
   H6P=h6p1,  prop_Eas_Asian=East_Asian,
   H7P=h7p1,  prop_European=European
)]
```

# Edit columns
## Winsorization, median imputation
```{r}
# Data may have duplicated individuals, e.g. because some individuals have multiple TOM_Ids.
# We want winsorization/imputation to be based on data without duplicated individuals,
# otherwise samples with more TOM_Ids would unfairly contribute more to quantiles, or medians, etc.
# because their phenotype measurements would appear more than once.
data_mesa_ids_unique <-
  data[
     # Deduplicate mesa_ids
     # Now there is one unique individual per row. No person has repeater phenotype measurements.
     ][ !duplicated(mesa_id)

     # Subset to only the phenotype cols we plan to edit. If we keep all columns, when merging later,
     # TOM_Ids would get overwritten by the one-per-sample that was arbitrarily chosen when we deduplicated mesa_id above.
     ][, .(mesa_id,
           pa, mod_pa, vig_pa, mvpa,
           ses_score, drinks_per_week, ahei_score, dash_score)

     # PA variables: divide by 60 then winsorize to 5 SDs
     ][, .SDcols=c('pa', 'mod_pa', 'vig_pa', 'mvpa'),
                 c('pa', 'mod_pa', 'vig_pa', 'mvpa') :=
          lapply(.SD, \(x) winsorizeBySd(x/60))

     # Select covariates: winsorize to 5 SDs then median-impute
     ][, .SDcols = c('ses_score','drinks_per_week','ahei_score','dash_score'),
                   c('ses_score','drinks_per_week','ahei_score','dash_score') :=
         lapply( .SD, \(x) { x <- winsorizeBySd(x); fifelse(is.na(x), median(x,na.rm=T), x) })
     ][, pa_bin := pa > 3.75
]

# The first df's columns take precedence when coalescing.
data <- mergecoalesce(data_mesa_ids_unique, data, all.y=T, by='mesa_id', force_coalesce=T)
```

## Conversion to factors, imputation, transformed PA variables.
```{r factorize-impute}
winsor      <- \(v, bounds=quantile(v,c(0,0.9),na.rm=T)) pmin(pmax(v,bounds[1]),bounds[2])
na_outliers <- \(v, bounds=quantile(v,c(0,0.9),na.rm=T)) fifelse(v<bounds[1] | v>bounds[2], NA, v)

data <- data[
  ][, hdl_log   := log(hdl) 
  ][, race      := fifelse(race1c==1, 'WHITE, CAUCASIAN',
                   fifelse(race1c==2, 'CHINESE AMERICAN',
                   fifelse(race1c==3, 'BLACK, AFRICAN-AMERICAN',
                   fifelse(race1c==4, 'HISPANIC', race1c)))) |>
                   as.factor()
  ][, site      := factor(site1c,  levels=3:8, labels=c('s3','s4','s5','s6','s7','s8'))
  ][, smoking   := factor(cig1c,   levels=0:2, labels=c('NEVER','FORMER','CURRENT')) ## There's also a 'Smoking Status' column but it has no additional info, and it's not clear what exam it's from so I'm sticking with cig1c.
  ][, income    := factor(income1, levels=c('< $5,000', '$5,000 - 7,999', '$8,000 - 11,999', '$12,000 - 15,999', '$16,000 - 19,999', '$20,000 - 24,999', '$25,000 - 29,999', '$30,000 - 34,999', '$35,000 - 39,999', '$40,000 - 49,999', '$50,000 - 74,999', '$75,000 - 99,999', '$100,000 +'))
  ][, education := factor(educ1,   levels=c('NO_SCHOOLING','GRADES 1-8','GRADES 9-11','COMPLETED HIGH SCHOOL/GED','SOME COLLEGE BUT NO DEGREE','ASSOCIATE DEGREE','TECHNICAL SCHOOL CERTIFICATE',"BACHELOR'S DEGREE",'GRADUATE OR PROFESSIONAL SCHOOL'))

  ][, sex := fcoalesce(gender1,Sex,SEX) |>
        sub(pattern='^F.*', replacement='FEMALE', ignore.case=T) |>
        sub(pattern='^M.*', replacement=  'MALE', ignore.case=T) |>
        factor(levels=c('FEMALE','MALE'))
      # There is an existing `sex` col in freeze9b_sample_annot_2020-08-20.txt,
      #   but it disagrees with the other `Sex`, `SEX`, and `gender1` cols. Ignoring it.

  # Imputation
  ][is.na(income),  income  := 'Missing'
  ][is.na(smoking), smoking := 'NEVER'

  # Transformed variables
  ][,     pa_trunc := na_outliers(    pa)
  ][,   mvpa_trunc := na_outliers(  mvpa)
  ][, mod_pa_trunc := na_outliers(mod_pa)
  ][, vig_pa_trunc := na_outliers(vig_pa)
  ][,     pa_wins  :=      winsor(    pa)
  ][,   mvpa_wins  :=      winsor(  mvpa)
  ][, mod_pa_wins  :=      winsor(mod_pa)
  ][, vig_pa_wins  :=      winsor(vig_pa)
  ][,     pa_log   :=         log(    pa+1)
  ][,   mvpa_log   :=         log(  mvpa+1)
  ][, mod_pa_log   :=         log(mod_pa+1)
  ][, vig_pa_log   :=         log(vig_pa+1)
]
```


# Append metabolomics PCs
We waited to calculate PCs until now because we only want to do PCA using samples we will use in the analysis; and now we are finally done filtering samples.\
We performed half-minimum imputation in metabolomics QC, but NAs are reintroduced when merging the metabolomics methods CP/CN/HP/AN, since not all samples got profiled with all the methods.\
PCA is intolerant to missing values, so we remove samples who have any missing values. Perhaps a more sophisticated imputation strategy would be better, but the mPCs are only used for exploratory analysis anyways.
```{r mPCA, cache=T}
met_nms <- names(fread('metabolomics/QCd/merged_QCd.csv',nrows=0,drop=1))
met_mtx <- data[
   ][ rowSums(is.na(data[,..met_nms]))==0 # Only samples with data for ALL metabolomics methods CP/CN/HP/AN, because PCA can't handle NAs
   ][ metabolomics_exam == 1
   ][, c('TOM_Id', ..met_nms)
] |> as.matrix(rownames=1)

pca <- prcomp(met_mtx, center=T, scale=T)
pcs <- as.data.table(pca$x[,1:20], keep.rownames=T)
setnames(pcs, c('TOM_Id',paste0('mPC', 1:20)))
data <- pcs[data, on='TOM_Id']

screeplot(pca,npcs=20)
```

## Write
```{r write}
fwrite(data, 'analysis_df.csv')
```
