```{css echo=F}
pre {
  overflow-x: scroll;
}
pre code {
  white-space: pre;
}
```

# Setup
```{r setup, message=F}
library(data.table)
library(readxl)

options(datatable.na.strings=c('NA',''))

if(!file.exists('phenotypes/')) {
  system('gsutil -m cp -r gs://fc-secure-4a392455-5587-4d6f-b8bd-01a1f834ae63/phenotypes .')
  system('gsutil    cp    gs://fc-secure-4b3e979d-ba8b-43c4-a5ec-b41ab42ce606/MesaMetabolomics_PilotX01_2023_02_16/MESA_PhenoUsmanSIDNO_20230214.txt phenotypes')
}

winsorizeBySd <- function(x, SDs=5) {
  lower_bound <- mean(x,na.rm=T) - SDs*sd(x,na.rm=T)
  upper_bound <- mean(x,na.rm=T) + SDs*sd(x,na.rm=T)
  print(paste(sum(x<lower_bound,na.rm=T), 'values winsorized at the lower bound.'))
  print(paste(sum(x>upper_bound,na.rm=T), 'values winsorized at the upper bound.'))
  x[x<lower_bound] <- lower_bound
  x[x>upper_bound] <- upper_bound
  return(x)
}
```

# Join all datasets
```{r join, cache=T}
mergecoalesce <- \(x,y,by, ..., force_coalesce=F) { # Merge which coalesces columns with duplicated names along the way
  setkeyv(x,by)
  setkeyv(y,by)
  merged <- merge(x,y, ...)

  nm_dups <- intersect(names(x),names(y)) |> setdiff(by)

  for(nm in nm_dups) {
    nms_w_suffix <- paste0(nm, c('.x','.y'))

    classes <- merged[, sapply(.SD,class), .SDcols=nms_w_suffix]
    if(!Reduce('==',classes)) {
      if(force_coalesce) { # Cast to character for maximum compatability
        merged[, (nms_w_suffix) := lapply(.SD, as.character), .SDcols = nms_w_suffix]
      } else {
        warning('Not coalescing these columns because their class doesn\'t match:\n', paste(capture.output(print(classes)), collapse='\n'))
        next
    }}

    merged <- merged[
        , (nm) := fcoalesce(.SD), .SDcols= nms_w_suffix 
      ][, .SD                   , .SDcols=!nms_w_suffix]
  }

  merged
}

data <- Reduce(\(x,l) mergecoalesce(x,l$y, by=l$by, all.x=T, force_coalesce=T),
    list(              x=fread('metabolomics/QCd/merged_QCd.csv'),
    list(by= 'TOM_Id', y=fread('metabolomics/sample_info.csv')                          [, TOM_Id := sample_id]),
    list(by= 'TOM_Id', y=fread('phenotypes/MESA_PhenoUsmanSIDNO_20230214.txt')          [, TOM_Id := tomid1][, mesa_id := sidno][, NWD_Id := fcoalesce(nwdid2,nwdid1)]), # OK to coalesce genotype
    list(by= 'TOM_Id', y=fread('phenotypes/MESA_PhenoUsmanSIDNO_20230214.txt')          [, TOM_Id := tomid2][, mesa_id := sidno][, NWD_Id := fcoalesce(nwdid2,nwdid1)]), #  info b/c it is not likely
    list(by= 'TOM_Id', y=fread('phenotypes/MESA_PhenoUsmanSIDNO_20230214.txt')          [, TOM_Id := tomid3][, mesa_id := sidno][, NWD_Id := fcoalesce(nwdid2,nwdid1)]), #  to change much across
    list(by= 'TOM_Id', y=fread('phenotypes/MESA_PhenoUsmanSIDNO_20230214.txt')          [, TOM_Id := tomid4][, mesa_id := sidno][, NWD_Id := fcoalesce(nwdid2,nwdid1)]), #  exams. But prefer the more
    list(by= 'TOM_Id', y=fread('phenotypes/MESA_PhenoUsmanSIDNO_20230214.txt')          [, TOM_Id := tomid5][, mesa_id := sidno][, NWD_Id := fcoalesce(nwdid2,nwdid1)]), #  recent genos from exam 2.
    list(by= 'TOM_Id', y=fread('phenotypes/id_match_file.csv')                          [, mesa_id := as.character(Cohort_Specific_Id)]),
    list(by='mesa_id', y=fread('phenotypes/SHARe_AncilMesaNMR_LP4_DS.txt')              [, mesa_id := as.character(subject_id)]),
    list(by='mesa_id', y=fread('phenotypes/mesa5_phenos_basic.csv')                     [, mesa_id := sub('phs000209.v13_', '', `\\_Parent Study Accession with Subject ID\\`)]),
    list(by='mesa_id', y=fread('phenotypes/mesa5_phenos_diet.csv')                      [, mesa_id := sub('phs000209.v13_', '', `\\_Parent Study Accession with Subject ID\\`)]),
    list(by='mesa_id', y=fread('phenotypes/primary_phenotypes.csv')                     [, mesa_id := sub('phs000209.v13_', '', `\\_Parent Study Accession with Subject ID\\`)]),
    list(by='mesa_id', y=fread('phenotypes/basic_phenotypes.csv')                       [, mesa_id := sub('phs000209.v13_', '', `\\_Parent Study Accession with Subject ID\\`)]),
    list(by='mesa_id', y=fread('phenotypes/nmr_metabolites.csv')                        [, mesa_id := sub('phs000209.v13_', '', `\\_Parent Study Accession with Subject ID\\`)]),
    list(by='mesa_id', y=fread('phenotypes/covariates.csv')                             [, mesa_id := sub('phs000209.v13_', '', `\\_Parent Study Accession with Subject ID\\`)]),
    list(by='mesa_id', y=fread('phenotypes/freeze9b_sample_annot_2020-08-20.txt')       [, mesa_id := subject_id][!duplicated(mesa_id)]),
    list(by='mesa_id', y=setDT(read_excel('phenotypes/draw_dates_MESA_study_site.xlsx'))[, mesa_id := sidno]),
    list(by= 'NWD_Id', y=fread('phenotypes/freeze9_pcair_results.tsv')                  [,  NWD_Id := sample.id]),
    list(by= 'NWD_Id', y=fread('genotypes.csv')                                         [,  NWD_Id := NWD_ID])
  )
)
```

# Rename columns
```{r rename}
setnames(data, \(nm) sub('.*\\\\(.*)\\\\$', '\\1', nm)) # Fixes some names e.g. '\\phs000209\\pht001116\\phv00084442\\age1c\\' -> 'age1c'
setnames(data, \(nm) sub('^PC'            , 'gPC', nm)) # 'PC#' -> 'gPC#' To distinguish genetic PCs from metabolite PCs we'll add later
data <- data[, `:=`(
   # SHARe...txt      #primary...csv    #nmr...csv           #covariates.csv           
   HDL_C=nhdlc1,      hdl=hdl1,         HDL_C_lp3=nhdlc31c,  ses_score=F1_PC2_1,
   HDL_P=chdlp1,      ldl=ldl1,         HDL_P_lp3= hdlp31c,  drinks_per_week=alcwkc1,
   HDL_size=hdlz1,    chol=chol1,       HDL_size_lp3=hz31,   ahei_score=ahei_2010_1,
   L_HDL_P=l_chdlp1,  tg=trig1,         L_HDL_P_lp3=hl31,    dash_score=dash_sodium1,
   M_HDL_P=m_chdlp1,      pa=exercm1c,  M_HDL_P_lp3=hm31,
   S_HDL_P=s_chdlp1,  mod_pa= pamcm1c,  S_HDL_P_lp3=hs31,
   H1P=h1p1,          vig_pa= pavcm1c,
   H2P=h2p1,            mvpa=pamvcm1c,
   H3P=h3p1,  #id_match_file.csv          #draw_dates...xlsx     #basic...csv                                              
   H4P=h4p1,  prop_African=African,       month=`exam1:month`,   age=age1c,
   H5P=h5p1,  prop_American=American,     season=`exam1:season`, bmi=bmi1c,
   H6P=h6p1,  prop_Eas_Asian=East_Asian,
   H7P=h7p1,  prop_European=European
)]
setnames(data, \(nm) gsub('\\\\','',nm)) # Rm backslashes from names
```

# Filter, and generate new columns
```{r filter}
data <- data[
  # Filtering
    is_qc_sample == F
  ][!is.na(pa) | !is.na(mod_pa) | !is.na(mvpa) | !is.na(vig_pa) | !is.na(rs295849)
  ][TOM_Id %in% tomid1 # Exam1 metabolomics only
  #][is.na(Exclusion_Reason)

  # Edit or creating custom columns
  ][, hdl_log   := log(hdl) 
  ][, race      := as.factor(race1c)
  ][, site      := factor(site1c,  levels=3:8, labels=c('s3','s4','s5','s6','s7','s8'))
  ][, smoking   := factor(cig1c,   levels=0:2, labels=c('NEVER','FORMER','CURRENT'))
  ][, income    := factor(income1, ordered=T,
                          levels=c('< $5,000', '$5,000 - 7,999', '$8,000 - 11,999', '$12,000 - 15,999', '$16,000 - 19,999', '$20,000 - 24,999', '$25,000 - 29,999', '$30,000 - 34,999', '$35,000 - 39,999', '$40,000 - 49,999', '$50,000 - 74,999', '$75,000 - 99,999', '$100,000 +'))
  ][, education := factor(educ1, ordered=T,
                          levels=c('NO_SCHOOLING','GRADES 1-8','GRADES 9-11','COMPLETED HIGH SCHOOL/GED','SOME COLLEGE BUT NO DEGREE','ASSOCIATE DEGREE','TECHNICAL SCHOOL CERTIFICATE',"BACHELOR'S DEGREE",'GRADUATE OR PROFESSIONAL SCHOOL'))

  ][, sex := fcoalesce(gender1,Sex,SEX) |>
        sub(pattern='^F.*', replacement='FEMALE', ignore.case=T) |>
        sub(pattern='^M.*', replacement=  'MALE', ignore.case=T) |>
        factor(levels=c('FEMALE','MALE'))
      # There is an existing `sex` col in freeze9b_sample_annot_2020-08-20.txt,\
      #   but it disagrees with the other `Sex`, `SEX`, and `gender1` cols.

  ## Some imputation of covariates to maintain sample size
  ][is.na(smoking), smoking := 'NEVER'

  ## Winsorization + median imputation
  ][, .SDcols = c('ses_score','drinks_per_week','ahei_score','dash_score'),
                c('ses_score','drinks_per_week','ahei_score','dash_score') :=
      lapply( .SD, \(x) { x <- winsorizeBySd(x); fifelse(is.na(x), median(x,na.rm=T), x) })

  ## Scaling + Winsorization
  ][, .SDcols=c('pa', 'mod_pa', 'vig_pa', 'mvpa'),
              c('pa', 'mod_pa', 'vig_pa', 'mvpa') :=
      lapply(.SD, \(x) winsorizeBySd(x/60))
  ][, pa_bin := as.numeric(pa > 3.75)

]
```

# Append metabolomics PCs
We waited to calculate PCs until now because we only want to do PCA using samples we will use in the analysis; and now we are finally done filtering samples.\
We performed half-minimum imputation in metabolomics QC, but NAs are reintroduced when merging the metabolomics methods CP/CN/HP/AN, since not all samples were profiled with all methods.\
PCA is intolerant to missing values, so we remove samples who have any missing values. Perhaps a more sophisticated imputation strategy would be better, but the mPCs are only used for exploratory analysis anyways.
```{r mPCA, cache=T}
met_nms <- names(fread('metabolomics/QCd/merged_QCd.csv',nrows=0,drop=1))
met_mtx <- data[ rowSums(is.na(data[,..met_nms]))==0, c('TOM_Id',..met_nms) ] |> as.matrix(rownames=1) # Only samples with data for ALL metabolomics methods
paste('Samples excluded from PCA:',sum(rowSums(is.na(data[,..met_nms]))>0),'/',nrow(data))

pca <- prcomp(met_mtx, center=T, scale=T)
pcs <- as.data.table(pca$x[,1:20], keep.rownames=T)
setnames(pcs, c('TOM_Id',paste0('mPC', 1:20)))
data <- pcs[data, on='TOM_Id']

screeplot(pca,npcs=20)
```

## Write
```{r write}
fwrite(data, 'analysis_df.csv')
```
