```{css echo=F}
pre {
  overflow-x: scroll;
}
pre code {
  white-space: pre;
}
```

# Setup
```{r setup, message=F}
library(data.table)
library(ggplot2)
library(gtsummary)
library(parallel)

options(datatable.na.strings=c('NA',''))
data <- fread('analysis_df.csv')
```

# Summary table of relevant variables
```{r summary-tbl, cache=T}
data[
  ][, c('bmi','race','smoking','income_cat','sex','pa','mod_pa','mvpa','vig_pa','age','ses_score','drinks_per_week','ahei_score','dash_score','rs295849')
  ][ is.na(sex), sex := 'Unknown'
  ] |>
  tbl_summary(
    by=sex,
    missing='no', # TODO more elegant way to display # of missing?
    statistic = list(
      all_continuous()  ~ c("{mean} ± {sd}"),
      all_categorical() ~ c("{n} ({p}%)"   )
  )) |>
  add_overall() |> add_n()
```

<!--
# Maybe mv those blu heatmaps from analysis to here
# Distributions of covariates and outcomes (hdl) and stuff
# Educ x race, and educ x income hists
-->

# Testing Sets of Covariates
`mPCs ~ e_or_snp + covars` for a few mPCs, all e's/snp, and all covar sets
```{r covar-sets, cache=T, fig.width=16, fig.height=10}
mPCs <- paste0('mPC',1:10)
vars2test <- c('rs295849','pa_bin','pa','mod_pa','mvpa','vig_pa')
covar_sets <- list(
  set1=c('site'                                                                                                                              ),
  set2=c('site', 'sex', 'age'                                                                                                        ),
  set3=c('site', 'sex', 'age', 'ses_score', 'income_cat'                                                                             ),
  set4=c('site', 'sex', 'age', 'ses_score', 'income_cat', 'drinks_per_week', 'smoking', 'ahei_score', 'dash_score'                   ),
  set5=c('site', 'sex', 'age', 'ses_score', 'income_cat', 'drinks_per_week', 'smoking', 'ahei_score', 'dash_score', 'race'           ),
  set6=c('site', 'sex', 'age', 'ses_score', 'income_cat', 'drinks_per_week', 'smoking', 'ahei_score', 'dash_score', paste0('gPC', 1:5))
)

runs <- CJ(mPCs, vars2test, covar_sets, sorted=F)[
  ][, lm_coefs := mcmapply(mPCs,vars2test,covar_sets, FUN = \(mPC,var,covar_set)
        lm(data[[mPC]] ~ ., data[,c(var,covar_set), with=F]) |> (\(model) summary(model)$coefficients)(),
      SIMPLIFY=F, mc.cores=4)

  # Extract results we want
  ][, var_coefs := mapply(lm_coefs, vars2test, FUN = \(coefs,var)
        coefs[rownames(coefs)==var,],
      SIMPLIFY=F)
  ][, estimate := sapply(var_coefs, '[', 'Estimate'  )
  ][, se       := sapply(var_coefs, '[', 'Std. Error')
  ][, p        := sapply(var_coefs, '[', 'Pr(>|t|)'  )
  ][, l95      := estimate-(1.96*se),
  ][, u95      := estimate+(1.96*se),

  # Cleanup
  ][, covar_sets := factor(covar_sets, levels=..covar_sets, labels=names(..covar_sets))
  ][, var_coefs := NULL
  ][, lm_coefs := NULL
]

ggplot(runs, aes(x=covar_sets, y=estimate, color=covar_sets)) +
  geom_point   (position=position_dodge(width=0.2)) +
  geom_errorbar(position=position_dodge(width=0.2), aes(ymin=l95,ymax=u95), width=0) +
  geom_hline(yintercept=0, color="gray") +
  scale_x_discrete(labels=NULL, breaks=NULL) + labs(x=NULL) +
  facet_grid(rows=vars(vars2test), cols=vars(mPCs), scales="free")
```

# Test Variable Transformations
HDL vs various transformations of mvpa as in `gxe-omics` workspace's `preliminary-analysis.ipynb`
```{r mvpa-tfs, cache=T}
winsor <- \(v, bounds=quantile(v,c(0.05,0.95),na.rm=T)) pmin(pmax(v,bounds[1]),bounds[2])

rbind(data[, .(hdl,        mvpa , tf='identity') ],
      data[, .(hdl,  trunc(mvpa),    'trunc'   ) ],
      data[, .(hdl,    log(mvpa),    'log'     ) ],
      data[, .(hdl, winsor(mvpa),    'winsor'  ) ], use.names=F) |>
  ggplot(aes(x=mvpa, y=hdl, group=tf)) +
  geom_smooth(method = 'gam', formula = y ~ s(x, bs = "cs"), se=T, na.rm=T) +
  facet_wrap(vars(tf), scales='free')
```





```{r}
library(tidyverse)
library(patchwork, table1)
```
```{r, include=F,eval=F}
theme_set(theme_bw())

## ggplot specifications ----------
ggplot_theme_standard_continuous <- theme_bw() + theme(
    axis.text.x = element_text(size=12, vjust=0.65, color = "black"),
    axis.text.y = element_text(size=12, color="black"), 
    strip.text=element_text(size=14, face="bold"),
    axis.title = element_text(size=14, color = "black")
)

ggplot_theme_standard_categorical <- theme_bw() + theme(
    axis.text.x = element_text(size=12, color = "black", angle=30, hjust=0.9),
    axis.text.y = element_text(size=12, color="black"), 
    strip.text=element_text(size=14, face="bold"),
    axis.title = element_text(size=14, color = "black")
)

## Color palettes
proj_palettes <- list(
    greens = paletteer::paletteer_dynamic("cartography::green.pal", 10),
    blues = paletteer::paletteer_dynamic("cartography::blue.pal", 10),
    oranges = paletteer::paletteer_dynamic("cartography::orange.pal", 10),
    pretty_dark = paletteer::paletteer_d("PrettyCols::Dark", 5))
  
```

## Build basic functions
Function to remove outliers by SD
```{r, include=F,eval=F}
remove_outliers.fun <- function(x, SDs=5) {
    bounds <- mean(x, na.rm=T) + SDs * c(-1, 1) * sd(x, na.rm=T)
    
    print(paste0(sum(x < bounds[1], na.rm=TRUE), " outliers removed at <", SDs, " SDs"))
    print(paste0(sum(x > bounds[2], na.rm=TRUE), " outliers removed at >", SDs, " SDs"))
    
    x <- ifelse(x>bounds[1] & x<bounds[2], x, NA) ; x    
}
```

Functions for describing data
```{r, include=F,eval=F}
# Print mean/sd
mean_sd<-function(x, d=2) {
  sprintf("%s \u00B1 %s", round(mean(x, na.rm=T), digits = d), 
          round(sd(x, na.rm=T), digits = d))
}

# Repeat from analysis_prep.Rmd
summarize_continuous <- function(x) {
  m <- round(mean(x, na.rm=TRUE), 1)
  s <- round(sd(x, na.rm=TRUE), 1)
  paste0(m, " (", s, ")")
}

# Print n_pct for categorical vars
n_pct <- function(x, level=F) {
  if(level==F) {
  sapply(as.list(names(table(x))), function(lvl) {
    paste0(lvl, ", ", sum(x == lvl, na.rm=T), " (", round(sum(x == lvl, na.rm=T)/n()*100,1), "%)") }) } 
  else{paste0(sum(x == level, na.rm=T), " (", round(sum(x == level, na.rm=T)/n()*100,1), "%)")}
}

```

```{r, include=F,eval=F}
# Cohort basic descriptives
pop_description_tbl <- analysis.df %>%
    group_by(RaceEthn) %>%
    reframe(
        N = n(),
        Age = mean_sd(age, d=1),
        Female = n_pct(female, level=1),
        BMI = mean_sd(bmi, d=1)) %>%
  arrange(desc(N))
pop_description_tbl %>% t()

# Repeat from analysis_prep.Rmd
pop_description_tbl <- analysis_df %>%
    rename(Race=race) %>%
  group_by(Race) %>%
  summarise(
    N = n(),
    Gender = paste(round(sum(gender_f0m1 == 0) / n() * 100, 1), "% female"),
    Age = summarize_continuous(age),
    BMI = summarize_continuous(bmi),
    `Intentional PA` = paste0(round(sum(pa_bin == 1, na.rm=TRUE) / n() * 100, 1), "%"),
    `Smoking (current)` = paste0(round(sum(smoking == "CURRENT", na.rm=TRUE) / n() * 100, 1), "%"),
    `Smoking (former)` = paste0(round(sum(smoking == "FORMER", na.rm=TRUE) / n() * 100, 1), "%"),
    `PCA-based measure of SES` = summarize_continuous(ses_score),
    
  ) %>%
  arrange(desc(N))
pop_description_tbl %>% t()
```

## Distributions
```{r, include=F,eval=F}
plot_continuous <- function(cont_var) {
  d_complete <- analysis.df %>% select(var=all_of(cont_var)) %>% filter(!is.na(var)) 
  d_complete %>% 
    ggplot(aes(x=var)) + geom_histogram(bins=30) +
    labs(title=cont_var, x=cont_var, y="frequency") +
    geom_vline(xintercept = mean(d_complete$var, na.rm=T), linewidth=1) +
    geom_vline(xintercept = c(mean(d_complete$var)+c(1,-1)*sd(d_complete$var)), linewidth=1, linetype="dashed") + 
    geom_vline(xintercept = median(d_complete$var, na.rm=T), linewidth=2, color = "red") +
    ggplot_theme_standard_continuous
}
plot_continuous <- function(cont_var) { # Repeat from analysis-prep.Rmd
    analysis_df %>%
    filter(!is.na(!!sym(cont_var))) %>%
    ggplot(aes_string(x=cont_var)) +
    geom_histogram(bins=30) +
    labs(title=cont_var)
}

plot_categorical <- function(cat_var) {
  analysis.df %>% 
    select(var=all_of(cat_var)) %>% filter(!is.na(var)) %>%
    ggplot(aes(x=factor(var))) + geom_bar(stat="count") +
    labs(title=cat_var, x=cat_var) +
    ggplot_theme_standard_categorical
}
plot_categorical <- function(cat_var) { # Repeat from analysis-prep.Rmd
    ggplot(analysis_df, aes(x=factor(!!sym(cat_var)))) +
    geom_bar(stat="count") +
    labs(title=cat_var) +
    theme(axis.text.x=element_text(angle=30, hjust=0.9))
}

plot_xyscatter <- function(x_var, y_var) {
  d_complete <- analysis.df %>% select(xvar=x_var, yvar = y_var) %>% filter(complete.cases(.)) 
  d_complete %>%
    ggplot(aes(x=xvar, y=yvar)) + geom_point(size=4, color="#00000075") + 
    labs(title=paste0(x_var, " by ", y_var, "\nr2 = ",
                      round(cor(d_complete$xvar, d_complete$yvar), 3))) +
    ylab(y_var) + xlab(x_var) +
    ggplot_theme_standard_continuous
}

options(repr.plot.width=14, repr.plot.height=5)
options(warn=-1)
```

### Basic descriptives 
```{r, include=F,eval=F}
# Basic descriptives
age_plt <- plot_continuous("age")
gender_plt <- plot_categorical("female")
bmi_plt <- plot_continuous("bmi")
age_plt + gender_plt + bmi_plt

racethn_plt <- plot_categorical("RaceEthn")
racethn_plt

# Repeat from analysis-prep
# Basic biological variables
age_plt <- plot_continuous("age")
gender_plt <- plot_categorical("gender_f0m1")
bmi_plt <- plot_continuous("bmi")
age_plt + gender_plt + bmi_plt

race_plt <- plot_categorical("race")
race_plt
```

### Plots from analysis-prep
```{r, include=F,eval=F}
# Outcomes
hdl_plt <- plot_continuous("hdl")
hdl_log_plt <- plot_continuous("hdl_log")
hdl_plt + hdl_log_plt

hdl_p_plt <- plot_continuous("HDL_P")
m_hdl_p_plt <- plot_continuous("M_HDL_P")
l_hdl_p_plt <- plot_continuous("L_HDL_P")
hdl_p_plt + m_hdl_p_plt + l_hdl_p_plt

# Exposures
pa_plt <- plot_continuous("pa")
pa_bin_plt <- plot_categorical("pa_bin")
pa_plt + pa_bin_plt

mvpa_plt <- plot_continuous("mvpa")
mod_pa_plt <- plot_continuous("mod_pa")
vig_pa_plt <- plot_continuous("vig_pa")
mvpa_plt + mod_pa_plt + vig_pa_plt

analysis_df %>%
    filter(!is.na(vig_pa)) %>%
    ggplot(aes(x=vig_pa, group=gender_f0m1, fill=factor(gender_f0m1))) +
#     geom_density() +
geom_histogram(bins=100) +
    labs(title="vig_pa by gender") +
coord_cartesian(ylim=c(0, 200))

# Batch variables
site_plt <- plot_categorical("site")
month_plt <- plot_categorical("month") + theme(axis.text.x=element_text(angle=30, hjust=0.9))
season_plt <- plot_categorical("season")
site_plt + month_plt + season_plt
```

### SES phenotypes
```{r, include=F,eval=F}
# Education & Income variables
educ_plt <- plot_categorical("educ_lvl.lab")
educ_4lvl_plt <- plot_categorical("educ_4lvl.lab")
educ_plt + educ_4lvl_plt

inc_plt <- plot_categorical("income_lvl.lab")
inc_4lvl_plt <- plot_categorical("income_4lvl.lab")
inc_plt + inc_4lvl_plt

# Somewhat Repeat from analysis-prep.Rmd
# Covariates
ses_plt <- plot_continuous("ses_score")
income_plt <- plot_categorical("income_cat")
ses_plt + income_plt

alc_plt <- plot_continuous("drinks_per_week")
smk_plt <- plot_categorical("smoking")
alc_plt + smk_plt

ahei_plt <- plot_continuous("ahei_score")
dash_plt <- plot_continuous("dash_score")
ahei_plt + dash_plt
```

#### Plot of education x race/ethnicity
```{r, include=F,eval=F}
# make plot for Education levels by R/E
as.data.frame(with(analysis.df, table(RaceEthn, educ_4lvl.lab))) %>% 
    ggplot(aes(x=RaceEthn, y=Freq, group=educ_4lvl.lab, fill=educ_4lvl.lab)) + 
    geom_bar(stat = "identity", position=position_dodge()) + 
    scale_fill_manual(values = c(proj_palettes$pretty_dark[1:5])) + 
    ggplot_theme_standard_categorical

as.data.frame(with(analysis.df, table(RaceEthn, educ_4lvl.lab))) %>% 
    ggplot(aes(x=educ_4lvl.lab, y=Freq, group=RaceEthn, fill=RaceEthn)) + 
    geom_bar(stat = "identity", position=position_dodge()) + 
    scale_fill_manual(values = c(proj_palettes$pretty_dark[1:5])) +
    ggplot_theme_standard_categorical
```

#### Plot of education x income
```{r, include=F,eval=F}
# Education x Income
as.data.frame(with(analysis.df, table(educ_4lvl.lab, income_4lvl.lab))) %>% 
    ggplot(aes(x=educ_4lvl.lab, y=Freq, group=income_4lvl.lab, fill=income_4lvl.lab)) + 
    geom_bar(stat = "identity", position=position_stack()) + 
    scale_fill_manual(values = c(proj_palettes$pretty_dark[1:5])) 
```

### Diet phenotypes
Total energy
```{r, include=F,eval=F}
energy_plt <- plot_continuous("energy_kcal")
energy_plt
```

Macronutrients (g & %kcal)
```{r, include=F,eval=F}
carb_plt <- plot_continuous("nut_carb_g")
carb_pct_plt <- plot_continuous("nut_carb_pct")
fat_plt <- plot_continuous("nut_fat_g")
fat_pct_plt <- plot_continuous("nut_fat_pct")
prot_plt <- plot_continuous("nut_prot_g")
prot_pct_plt <- plot_continuous("nut_prot_pct")
carb_plt + carb_pct_plt 
fat_plt + fat_pct_plt
prot_plt + prot_pct_plt
```

Fat types (g & %kcal)
```{r, include=F,eval=F}
mufa_plt <- plot_continuous("nut_mufa_g")
mufa_pct_plt <- plot_continuous("nut_mufa_pct")
pufa_plt <- plot_continuous("nut_pufa_g")
pufa_pct_plt <- plot_continuous("nut_pufa_pct")
sfa_plt <- plot_continuous("nut_sfa_g")
sfa_pct_plt <- plot_continuous("nut_sfa_pct")

mufa_plt + mufa_pct_plt
pufa_plt + pufa_pct_plt
sfa_plt + sfa_pct_plt
```

### Carbohydrate quality
```{r, include=F,eval=F}
fib_plt <- plot_continuous("nut_fiber_g")
fibsol_plt <- plot_continuous("nut_fiber_sol_g")
fibinsol_plt <- plot_continuous("nut_fiber_insol_g")
fib_plt 
fibsol_plt + fibinsol_plt

# note: soluble + insoluble = total
```

**Carbohydrate-to-fiber ratio
```{r, include=F,eval=F}
carb2fib_plt <- plot_continuous("carb2fib") 
fib2carb_plt <- plot_continuous("fib2carb") 
carb2fib_plt + fib2carb_plt
```

```{r, include=F,eval=F}
## Add ranges, means/sd, medians
c("Carb-to-Fiber: ", round(quantile(analysis.df$carb2fib, na.rm=T, probs=seq(0,1,0.2), include.lowest=T), 1))
c("Fiber-to-Carb: ", round(quantile(analysis.df$fib2carb, na.rm=T, probs=seq(0,1,0.2), include.lowest=T), 3))

# Create boxplot to identify potential outliers
carb2fib_boxplt <- analysis.df %>% select("carb2fib") %>%
    filter(!is.na("carb2fib")) %>%
    ggplot(aes(x=carb2fib)) + 
    geom_boxplot(outlier.color="red") +
    ggplot_theme_standard_continuous

# Create boxplot to identify potential outliers
fib2carb_boxplt <- analysis.df %>% select("fib2carb") %>%
    filter(!is.na("fib2carb")) %>%
    ggplot(aes(x=fib2carb)) + 
    geom_boxplot(outlier.color="red") +
    ggplot_theme_standard_continuous

carb2fib_boxplt
fib2carb_boxplt
```

```{r, include=F,eval=F}
#carb-to-fib
bounds_5SD <- c(mean(analysis.df$carb2fib, na.rm=T)+c(-5,5)*sd(analysis.df$carb2fib, na.rm=T))
bounds_5to95 <- quantile(analysis.df$carb2fib, na.rm=T, probs = c(0.05, 0.25, 0.5, 0.75, 0.95))

analysis.df %>% 
select("carb2fib") %>%
    filter(!is.na("carb2fib")) %>%
    mutate(gt5SD = ifelse(carb2fib < bounds_5SD[1] | carb2fib > bounds_5SD[2],">5 SD", "ok")) %>%
    mutate(out5to95 = ifelse(carb2fib < bounds_5to95[1] | carb2fib > bounds_5to95[5],">5 to 95%", "ok")) %>%
    reframe(Outliers_carb2fib_gt5SD=n_pct(gt5SD),
           Outliers_carb2fib_out5to95=n_pct(out5to95))


#fib-to-carb
bounds_5SD <- c(mean(analysis.df$fib2carb, na.rm=T)+c(-5,5)*sd(analysis.df$fib2carb, na.rm=T))
bounds_5to95 <- quantile(analysis.df$fib2carb, na.rm=T, probs = c(0.05, 0.25, 0.5, 0.75, 0.95))

analysis.df %>% 
select("fib2carb") %>%
    filter(!is.na("fib2carb")) %>%
    mutate(gt5SD = ifelse(fib2carb < bounds_5SD[1] | fib2carb > bounds_5SD[2],">5 SD", "ok")) %>%
    mutate(out5to95 = ifelse(fib2carb < bounds_5to95[1] | fib2carb > bounds_5to95[5],">5 to 95%", "ok")) %>%
    reframe(Outliers_fib2carb_gt5SD=n_pct(gt5SD),
           Outliers_fib2carb_out5to95=n_pct(out5to95))

```

Note on outliers for Carbohydrate-to-fiber ratio: 
* 13 (0.3%) values are outside mean+/- 5 SDs
* 406 (10%) vauues are outside 5 to 95% of the data

### Covariates
Lifestyle factors
```{r, include=F,eval=F}
smk_plt <- plot_categorical("smoke_stat.lab")
smkyrs_plt <- plot_continuous("smoke_packyrs")
alc_plt <- plot_categorical("alch_currdrnk.lab")
pa_plt <- plot_continuous("physact_mvpa")
smk_plt + smkyrs_plt
alc_plt + pa_plt
```

Health status-related covariates
```{r, include=F,eval=F}
med_t2d_plt <- plot_categorical("med_t2d")
med_htn_plt <- plot_categorical("med_htn")
med_lip_plt <- plot_categorical("med_lip")
med_t2d_plt + med_htn_plt + med_lip_plt
```

```{r, include=F,eval=F}
missingness_vars <- c(
    "age", "gender", "bmi", "RaceEthn",
    "educ_4lvl.lab", "income_4lvl.lab",
    "energy_kcal",
    "nut_carb_g", "carb2fib", "nut_fat_g", "nut_prot_g",
    "alch_currdrnk.lab", "smoke_stat.lab",
    "med_t2d", "med_htn", "med_lip")

analysis.df %>%
    select(all_of(missingness_vars)) %>%
    mutate(across(everything(), is.na)) %>%
    summarise(across(everything(), sum)) %>% t()
```

```{r, include=F,eval=F}
# Re-run basic participant descriptives

analysis.df %>%
    group_by(RaceEthn) %>%
    reframe(
        N = n(),
        Age = mean_sd(age, d=1),
        Female = n_pct(female, level=1),
        BMI = mean_sd(bmi, d=1)) %>%
      arrange(desc(N)) %>% t()
```

# Run basic correlations among diet variables
### Carbohydrate & macronutrients
```{r, include=F,eval=F}
options(repr.plot.width=14, repr.plot.height=5)
options(warn=-1)

# Carbohydrates vs. macros
carbfat_plt<-plot_xyscatter("nut_carb_g", "nut_fat_g")
carbprot_plt<-plot_xyscatter("nut_carb_g", "nut_prot_g")
carbengy_plot <- plot_xyscatter("nut_carb_g", "energy_kcal")
carbfat_plt + carbprot_plt + carbengy_plot

# Carbohydrates vs. types of fat
carbmufa_plt<-plot_xyscatter("nut_carb_g", "nut_mufa_g")
carbpufa_plt<-plot_xyscatter("nut_carb_g", "nut_pufa_g")
carbsfa_plot <- plot_xyscatter("nut_carb_g", "nut_sfa_g")
carbmufa_plt + carbpufa_plt + carbsfa_plot
```

### Carbohydrate & fiber sources
```{r, include=F,eval=F}
carb2fib_plt <- plot_continuous("carb2fib")
carbfib_plt<-plot_xyscatter("nut_carb_g", "nut_fiber_g")
carb2fib_plt + carbfib_plt 

carbfibsol_plt<-plot_xyscatter("nut_carb_g", "nut_fiber_sol_g")
carbfibinsol_plt<-plot_xyscatter("nut_carb_g", "nut_fiber_insol_g")
carbfibsol_plt + carbfibinsol_plt + carbfibcer_plt
```

# Covariate descriptions across education levels
### Demographic, behavioral & lifestyle phenotypes
```{r, include=F,eval=F}
analysis.df %>% 
    group_by(educ_4lvl.lab) %>%
    reframe(
        N = n(),
        Age = mean_sd(age, d=1),
        Female = n_pct(female, level=1),
        BMI = mean_sd(bmi, d=1),
        Smoking_Current = n_pct(smoke_stat.lab, level="Current"),
        Smoking_Former_lt1yr = n_pct(smoke_stat.lab, level="Former, <1 yr"),
        Smoking_Former_gt1yr = n_pct(smoke_stat.lab, level="Former, >1 yr"),
        Smoking_Never = n_pct(smoke_stat.lab, level="Never"),
        PA_MVPA = mean_sd(physact_mvpa),
        Alchohol_Drinker = n_pct(alch_currdrnk.lab, level="Drinker"),
        Medication_t2d = n_pct(med_t2d, level=1),
        Medication_htn = n_pct(med_htn, level=1),
        Medication_lipid = n_pct(med_lip, level=1),
        Fasting_Glucose= mean_sd(glucose),
        HbA1c=mean_sd(hba1c),
        Triglyceride=mean_sd(tg),
        LDL=mean_sd(ldl),
        HDL=mean_sd(hdl)
    ) %>% t()
```

### Diet phenotypes
Macronutrients
```{r, include=F,eval=F}
analysis.df %>% 
    group_by(educ_4lvl.lab) %>%
    reframe(
        N = n(),
        Energy_kcal=mean_sd(energy_kcal),
        Carb_g=mean_sd(nut_carb_g),
        Protein_g=mean_sd(nut_prot_g),
        Fat_g=mean_sd(nut_fat_g),
        MUFA_g=mean_sd(nut_mufa_g),
        PUFA_g=mean_sd(nut_pufa_g),
        SFA_g=mean_sd(nut_sfa_g),
        Carb_pct=mean_sd(nut_carb_pct),
        Protein_pct=mean_sd(nut_prot_pct),
        Fat_pct=mean_sd(nut_fat_pct),
        MUFA_pct=mean_sd(nut_mufa_pct),
        PUFA_pct=mean_sd(nut_pufa_pct),
        SFA_pct=mean_sd(nut_sfa_pct)
    ) %>% t()
```

Carbohydrate Quality
```{r, include=F,eval=F}
analysis.df %>% 
    group_by(educ_4lvl.lab) %>%
    reframe(
        N = n(),
        Carb_to_Fiber=mean_sd(carb2fib),
        Fiber_to_Carb=mean_sd(fib2carb, d=3),
        Fiber_g=mean_sd(nut_fiber_g),
        Fiber_soluble_g=mean_sd(nut_fiber_sol_g),
        Fiber_insoluble_g=mean_sd(nut_fiber_insol_g)
    ) %>% t()
```



# BELOW IS WHAT OTHER STUFF THERE WAS LEFT IN ANALYSIS-PREP.ipynb


# Covariate associations with metabolomic PCs
```{r, include=F,eval=F}
race_plt <- bind_cols(analysis_df_lcms, pc_res$x[, 1:5]) %>%
    ggplot(aes(x=PC1, y=PC2)) +
    geom_point(aes(color=race), alpha=0.5)

site_plt <- bind_cols(analysis_df_lcms, pc_res$x[, 1:5]) %>%
    ggplot(aes(x=PC1, y=PC2)) +
    geom_point(aes(color=site), alpha=0.5)

gender_plt <- bind_cols(analysis_df_lcms, pc_res$x[, 1:5]) %>%
    ggplot(aes(x=PC1, y=PC2)) +
    geom_point(aes(color=gender1), alpha=0.5)

options(repr.plot.width=12, repr.plot.height=3)

race_plt + site_plt + gender_plt
```

```{r, include=F,eval=F}
cont_vars <- c("age", "bmi", 
#                paste0("prop_", ancestry_prop_fields), 
               paste0("gPC", 1:11))
bin_vars <- c("smoking", "gender_f0m1")
cat_vars <- c("race", "site", "season", "month")

pc_test <- function(pc_var, covariate) {
    lm_res <- lm(pc_res$x[, pc_var] ~ analysis_df_lcms[[covariate]])
    model_p <- anova(lm_res)[1, 5]
    model_p
}

pc_test_res <- expand_grid(
    pc = paste0("PC", 1:20),
    covar = c(cont_vars, bin_vars, cat_vars)
) %>%
    rowwise() %>%
    mutate(p = pc_test(pc, covar))
```

```{r, include=F,eval=F}
pc_test_plt_df <- pc_test_res %>%
    mutate(nlp = -log10(p),
           sig = ifelse(p < 0.05, "*", ""),
           pc = factor(pc, levels=paste0("PC", 1:20)))

options(repr.plot.width=12, repr.plot.height=6)

pc_test_plt_df %>%
    ggplot(aes(x=pc, y=covar, fill=nlp)) +
    geom_tile() +
    geom_text(aes(label=sig), color="white") +
    scale_fill_continuous(name=expression(-log[10] * "(p)")) +
    scale_size_area(max_size=1.5) +
    labs(title="Covariate-mPC association p-values")

pc_test_plt_df %>%
    mutate(nlp=pmin(nlp, 10)) %>%
    ggplot(aes(x=pc, y=covar, fill=nlp)) +
    geom_tile() +
    geom_text(aes(label=sig), color="white") +
    scale_fill_continuous(name=expression(-log[10] * "(p)")) +
    scale_size_area(max_size=1.5) +
    labs(title="Covariate-mPC association p-values (capped below p<1e-10)")
```

# Assessment of the SNPs and exposures of interest
```{r, include=F,eval=F}
snp_info_df <- read_csv("genotypes/snp_info.csv", col_types=cols())
exposures <- c("pa_bin")
```

## Univariate SNP-mPC and exposure-mPC associations
```{r, include=F,eval=F}
snp_mPC_assoc_df <- lapply(snp_info_df$rsID, function(rsID) {
    g_dominant <- analysis_df_lcms[[rsID]] > 0
    form_str <- paste0("g_dominant ~ ", paste0("mPC", 1:20, collapse=" + "))
    glm(as.formula(form_str), data=analysis_df_lcms, family="binomial") %>%
        broom::tidy()
}) %>%
    setNames(snp_info_df$rsID) %>%
    bind_rows(.id="snp")

exp_mPC_assoc_df <- lapply(exposures, function(e) {
    e <- analysis_df_lcms[[e]]
    form_str <- paste0("e ~ ", paste0("mPC", 1:20, collapse=" + "))
    glm(as.formula(form_str), data=analysis_df_lcms, family="binomial") %>%
        broom::tidy()
}) %>%
    setNames(exposures) %>%
    bind_rows(.id="exposure")
```

```{r, include=F,eval=F}
snp_mPC_assoc_df %>%
    filter(term != "(Intercept)") %>%
    mutate(nlp = -log10(p.value),
           term = factor(term, levels=paste0("mPC", 1:20))) %>%
    ggplot(aes(x=term, y=snp, fill=nlp)) +
    geom_tile() +
    scale_fill_continuous(name=expression(-log[10] * "(p)"))

exp_mPC_assoc_df %>%
    filter(term != "(Intercept)") %>%
    mutate(nlp = -log10(p.value),
           term = factor(term, levels=paste0("mPC", 1:20))) %>%
    ggplot(aes(x=term, y=exposure, fill=nlp)) +
    geom_tile() +
    scale_fill_continuous(name=expression(-log[10] * "(p)"))
```


TODO: try to squish the "Incorporation of technical" AND "biological" covariates sections down into one?
## Incorporation of technical covariates
```{r, include=F,eval=F}
technical_covar_sets <- list(
    none = "",
    add_site = c("gender_f0m1", "age", "site")
)

test_univariate <- function(y, x, covar_vec) {
    form_str <- paste0(y, " ~ ", x)
    if (!identical(covar_vec, "")) {
        form_str <- paste0(form_str, " + ", paste(covar_vec, collapse=" + "))
    }
    lm(as.formula(form_str), data=analysis_df_lcms) %>%
        broom::tidy() %>%
        filter(term == x)
}
```

```{r, include=F,eval=F}
snp_mPC_technical_covar_assoc_df <- expand_grid(
    snp = snp_info_df$rsID,
    mPC = paste0("mPC", 1:3),
    covar_set = names(technical_covar_sets) 
) %>%
    rowwise() %>%
    mutate(lm_res = list(test_univariate(mPC, snp, technical_covar_sets[[covar_set]]))) %>%
    unnest(lm_res)

exp_mPC_technical_covar_assoc_df <- expand_grid(
    e = exposures,
    mPC = paste0("mPC", 1:3),
    covar_set = names(technical_covar_sets) 
) %>%
    rowwise() %>%
    mutate(lm_res = list(test_univariate(mPC, e, technical_covar_sets[[covar_set]]))) %>%
    unnest(lm_res)
```

```{r, include=F,eval=F}
snp_mPC_technical_covar_assoc_df %>%
    mutate(l95 = estimate - 1.96 * std.error,
           u95 = estimate + 1.96 * std.error,
           covar_set = factor(covar_set, levels=names(technical_covar_sets))) %>%
    ggplot(aes(x=snp, y=estimate, color=covar_set)) +
    geom_point(position=position_dodge(width=0.2)) +
    geom_errorbar(aes(ymin=l95, ymax=u95), 
                  position=position_dodge(width=0.2), width=0.1) +
    geom_hline(yintercept=0, color="gray") +
    facet_wrap(~mPC, ncol=1, scales="free")

exp_mPC_technical_covar_assoc_df %>%
    mutate(l95 = estimate - 1.96 * std.error,
           u95 = estimate + 1.96 * std.error,
           covar_set = factor(covar_set, levels=names(technical_covar_sets))) %>%
    ggplot(aes(x=e, y=estimate, color=covar_set)) +
    geom_point(position=position_dodge(width=0.2)) +
    geom_errorbar(aes(ymin=l95, ymax=u95), 
                  position=position_dodge(width=0.2), width=0.1) +
    geom_hline(yintercept=0, color="gray") +
    facet_wrap(~mPC, ncol=1, scales="free")
```

## Incorporation of biological covariates
```{r, include=F,eval=F}
biological_covar_sets <- list(
    basic                   =   "site",
    add_gender_age          = c("site", "gender_f0m1", "age"),
    add_gender_age_race     = c("site", "gender_f0m1", "age", "race"),
    add_gender_age_ses      = c("site", "gender_f0m1", "age", "ses_score", "income_cat"),
    add_gender_age_ses_HL   = c("site", "gender_f0m1", "age", "ses_score", "income_cat", 
                              "drinks_per_week", "smoking", "ahei_score", "dash_score"),
    add_gender_age_gPC      = c("gender_f0m1", "age", paste0("gPC", 1:5)),
    add_gender_age_race_gPC = c("site", "gender_f0m1", "age", "race", paste0("gPC", 1:5))
)
```

```{r, include=F,eval=F}
snp_mPC_biological_covar_assoc_df <- expand_grid(
    snp = snp_info_df$rsID,
    mPC = paste0("mPC", 1:3),
    covar_set = names(biological_covar_sets) 
) %>%
    rowwise() %>%
    mutate(lm_res = list(test_univariate(snp, mPC, biological_covar_sets[[covar_set]]))) %>%
    unnest(lm_res)

exp_mPC_biological_covar_assoc_df <- expand_grid(
    e = exposures,
    mPC = paste0("mPC", 1:3),
    covar_set = names(biological_covar_sets) 
) %>%
    rowwise() %>%
    mutate(lm_res = list(test_univariate(e, mPC, biological_covar_sets[[covar_set]]))) %>%
    unnest(lm_res)
```

```{r, include=F,eval=F}
snp_mPC_biological_covar_assoc_df %>%
    mutate(l95 = estimate - 1.96 * std.error,
           u95 = estimate + 1.96 * std.error,
           covar_set = factor(covar_set, levels=names(biological_covar_sets))) %>%
    ggplot(aes(x=snp, y=estimate, color=covar_set)) +
    geom_point(position=position_dodge(width=0.2)) +
    geom_errorbar(aes(ymin=l95, ymax=u95), 
                  position=position_dodge(width=0.2), width=0.1) +
    geom_hline(yintercept=0, color="gray") +
    facet_wrap(~mPC, ncol=1, scales="free")

exp_mPC_biological_covar_assoc_df %>%
    mutate(l95 = estimate - 1.96 * std.error,
           u95 = estimate + 1.96 * std.error,
           covar_set = factor(covar_set, levels=names(biological_covar_sets))) %>%
    ggplot(aes(x=e, y=estimate, color=covar_set)) +
    geom_point(position=position_dodge(width=0.2)) +
    geom_errorbar(aes(ymin=l95, ymax=u95), 
                  position=position_dodge(width=0.2), width=0.1) +
    geom_hline(yintercept=0, color="gray") +
    facet_wrap(~mPC, ncol=1, scales="free")
```

# Conclusions
* The most important covariates affecting top PCs are study site and race (highly correlated) as well as gender
* It appears that most of the high-level metabolite associations can be captured with about 9 metabolite PCs (for both genotypes and exposures) 
* So, we want to adjust for PEER factors, and we are OK knowing that they are representing expected variables.
# Export final datasets for analysis
```{r, include=F,eval=F}
write_csv(analysis_df, "analysis/analysis_df.csv")  # All individuals
write_csv(analysis_df_lcms, "analysis/analysis_df_lcms.csv")  # Matched to the LC/MS dataset
saveRDS(mesa_metab_mat, "metabolites/lcms_metabolites.rds")
system(paste0("gsutil cp -R phenotypes analysis metabolites $WORKSPACE_BUCKET/"))
```
