```{css echo=F}
pre {
  overflow-x: scroll;
}
pre code {
  white-space: pre;
}
```

# Setup
```{r setup, message=F, cache=T, cache.depends=file.mtime('analysis_df.csv')}
library(circlize)
library(ComplexHeatmap)
library(data.table)
library(ggplot2)
library(gtsummary)
library(parallel)

options(datatable.na.strings=c('NA',''))
data <- fread('analysis_df.csv')[,exam:=as.factor(exam)]
```

# Summary table of relevant variables
```{r summary-tbl, cache=T, cache.depend=digest::digest(data)}
data[
  ][, .(age, bmi, sex, pa, mod_pa, mvpa, vig_pa, ahei_score, dash_score, ses_score, drinks_per_week, rs295849)
  ][, has_metabolomics := !is.na(data$TOM_Id)
  ][ is.na(sex), sex := 'Unknown'
  ] |>
  tbl_summary(
    by=sex,
    missing='no', # TODO more elegant way to display # of missing?
    statistic = list(
      all_continuous()  ~ c("{mean} ± {sd}"),
      all_categorical() ~ c("{n} ({p}%)"   )
  )) |>
  add_overall() |> add_n()
```

# Heatmap of mPCs vs. covars
Note that not all samples contributed to the mPCs — only samples with measurements for all metabolomics methods CP/CN/HP/AN could be used in PCA.
```{r hm, cache=T, cache.depend=digest::digest(data)}
heatmap_covars <- c(
  paste0('gPC', 1:11),
  'site', 'season', 'month',
  'age', 'bmi', 'sex', 'race',
  'ahei_score', 'dash_score', 'ses_score', 
  'income', 'drinks_per_week', 'smoking')

runs <- CJ(heatmap_covars, mPC=paste0('mPC',1:20), sorted=F)[
  ][, lm_res    := mapply(mPC,heatmap_covars, FUN = \(mPC,covar) lm(data[[mPC]] ~ data[[covar]]))
  ][, anova_res := lapply(lm_res, anova)
  ][, p         := sapply(anova_res, \(res) res$`Pr(>F)`[1])
]
p_mtx <- as.matrix(dcast(runs, heatmap_covars~mPC, value.var='p'), rownames=1)
p_mtx <- p_mtx[,paste0('mPC',1:20)]

# TODO also do other way of doing heatmap, see journal day 241112.
# Covars, as above: lm(mPC ~ one_covar)
# SNPs, glm(snp ~ all_mPCs, family='binomial')

Heatmap(-log10(p_mtx),
  cell_fun = \(j,i,x,y,w,h,col) if(p_mtx[i,j]<0.05) grid.text('*',x,y,gp=gpar(col='white')),
  col = colorRamp2(c(0,10), c('black','steelblue')),
  cluster_columns=F, cluster_rows=F,
  heatmap_legend_param = list(title = 'log10(p)\nfrom lm')
)
```

Define covariate sets we'll want to test
```{r covar-sets, cache=T, cache.depend=digest::digest(data)}
covar_sets <- list(
  site                    =c('site'                                                                                                                           ),
  `+sex`                  =c('site', 'sex'                                                                                                                    ),
  `+age`                  =c('site', 'sex', 'age'                                                                                                             ),
  `+ses+income`           =c('site', 'sex', 'age', 'ses_score', 'income'                                                                                      ),
  `+drink+smoke+ahei+dash`=c('site', 'sex', 'age', 'ses_score', 'income', 'drinks_per_week', 'smoking', 'ahei_score', 'dash_score'                            ),
  `+race`                 =c('site', 'sex', 'age', 'ses_score', 'income', 'drinks_per_week', 'smoking', 'ahei_score', 'dash_score', 'race'                    ),
  `+gPCs`                 =c('site', 'sex', 'age', 'ses_score', 'income', 'drinks_per_week', 'smoking', 'ahei_score', 'dash_score', 'race', paste0('gPC', 1:5)),
  `-race`                 =c('site', 'sex', 'age', 'ses_score', 'income', 'drinks_per_week', 'smoking', 'ahei_score', 'dash_score',         paste0('gPC', 1:5)))
```

# Inspect Variable Transformations
HDL vs various transformations of mvpa as in `gxe-omics` workspace's `preliminary-analysis.ipynb`
```{r mvpa-tfs, cache=2, cache.depend=digest::digest(data), fig.height=14, fig.width=10}
tmp <- data[exam==1]
tmp <- tmp[
  ][, mvpa_resid             := resid(lm(mvpa       ~ ., tmp[, .SD, .SDcols=        covar_sets$`+gPCs`       ], na.action=na.exclude))
  ][, mvpa_log_resid         := resid(lm(mvpa_log   ~ ., tmp[, .SD, .SDcols=        covar_sets$`+gPCs`       ], na.action=na.exclude))
  ][, mvpa_wins_resid        := resid(lm(mvpa_wins  ~ ., tmp[, .SD, .SDcols=        covar_sets$`+gPCs`       ], na.action=na.exclude))
  ][, mvpa_trunc_resid       := resid(lm(mvpa_trunc ~ ., tmp[, .SD, .SDcols=        covar_sets$`+gPCs`       ], na.action=na.exclude))
  ][, mvpa_resid_nosex       := resid(lm(mvpa       ~ ., tmp[, .SD, .SDcols=setdiff(covar_sets$`+gPCs`,'sex')], na.action=na.exclude))
  ][, mvpa_log_resid_nosex   := resid(lm(mvpa_log   ~ ., tmp[, .SD, .SDcols=setdiff(covar_sets$`+gPCs`,'sex')], na.action=na.exclude))
  ][, mvpa_wins_resid_nosex  := resid(lm(mvpa_wins  ~ ., tmp[, .SD, .SDcols=setdiff(covar_sets$`+gPCs`,'sex')], na.action=na.exclude))
  ][, mvpa_trunc_resid_nosex := resid(lm(mvpa_trunc ~ ., tmp[, .SD, .SDcols=setdiff(covar_sets$`+gPCs`,'sex')], na.action=na.exclude))
]

rbind(tmp[, .(sex, hdl_log, d1=mvpa,            d2=mvpa,                 f='identity'   ) ],
      tmp[, .(sex, hdl_log,    mvpa_resid,         mvpa_resid_nosex,       'resid'      ) ],
      tmp[, .(sex, hdl_log,    mvpa_log,           mvpa_log,               'log'        ) ],
      tmp[, .(sex, hdl_log,    mvpa_log_resid,     mvpa_log_resid_nosex,   'resid_log'  ) ],
      tmp[, .(sex, hdl_log,    mvpa_wins,          mvpa_wins,              'winsor'     ) ],
      tmp[, .(sex, hdl_log,    mvpa_wins_resid,    mvpa_wins_resid_nosex,  'resid_wins' ) ],
      tmp[, .(sex, hdl_log,    mvpa_trunc,         mvpa_trunc,             'trunc'      ) ],
      tmp[, .(sex, hdl_log,    mvpa_trunc_resid,   mvpa_trunc_resid_nosex, 'resid_trunc') ],
      use.names=F)[
    ][, f := factor(f,levels=unique(f)) ] |>
  ggplot(aes(y=hdl_log)) +
  geom_smooth(aes(x=d1           ), method = 'gam', formula = y ~ s(x, bs='cs'), se=T, na.rm=T) +
  geom_smooth(aes(x=d2, color=sex), method = 'gam', formula = y ~ s(x, bs='cs'), se=T, na.rm=T) +
  facet_wrap(vars(f), ncol=2, scales='free') +
  ggtitle(label='Residuals from `mvpa_variable ~ covars_except_sex`')
```

# Run models
Strategy: run all lm models upfront and examine the results afterward.\
One set of models we want to test is `mPC ~ pa_or_snp_variable + covar_set` for each possible combination of mPC1-10, untransformed PA or SNP variable, and covariate set. The point of this set of models is to see which covariates make the largest difference in the relationship between our variables of interest and the metabolomics, to help figure out which covariates to adjust for.\
Another set of models is `hdl_log ~ pa_or_snp_variable_transformed + covar_set`, for each possible combination of PA or SNP variable, and covariate set. The point of this is to see the main effects, and also see which covariates make the largest difference in these main effects.
```{r lms, cache=T, cache.depend=digest::digest(data)}
# mPC ~ pa_or_snp + covars
ys <- paste0('mPC',1:9)
xs <- c('rs295849','pa_bin','pa','mod_pa','mvpa','vig_pa')
lm_runs_w_mPC_outcome <- CJ(ys, xs, covar_sets, sorted=F)

# hdl_log ~ transformed_pa_or_snp + covars
ys <- 'hdl_log'
xs <- c('rs295849', 'pa',       'mod_pa',       'mvpa',       'vig_pa',
                    'pa_log',   'mod_pa_log',   'mvpa_log',   'vig_pa_log',
                    'pa_wins',  'mod_pa_wins',  'mvpa_wins',  'vig_pa_wins',
                    'pa_trunc', 'mod_pa_trunc', 'mvpa_trunc', 'vig_pa_trunc')
lm_runs_w_hdl_outcome <- CJ(ys, xs, covar_sets, sorted=F)

runs <- rbind(lm_runs_w_mPC_outcome,
              lm_runs_w_hdl_outcome)
runs <- runs[, fmlas := paste0('scale(',ys,')','~','scale(',xs,')','+',sapply(covar_sets,paste,collapse='+')) |> sapply(formula) ]

runs[
  ][, lm_coefs := mcmapply(fmlas, FUN = \(fmla)
        lm(fmla,data[exam==1]) |> (\(model) summary(model)$coefficients)(),
      SIMPLIFY=F,     # ↑ Pipe into a function to extract only the coefs. Don't need
      mc.cores=4)     #   all of lm's output. Keeping everything eats up a lot of memory.
  # Extract the results we want
  ][, var_coefs := mapply(lm_coefs, xs, FUN = \(coefs,x)
        coefs[grepl(x,rownames(coefs)),],
      SIMPLIFY=F)

  ][, estimate := sapply(var_coefs, '[', 'Estimate'  )
  ][, se       := sapply(var_coefs, '[', 'Std. Error')
  ][, t        := sapply(var_coefs, '[', 't value'   )
  ][, p        := sapply(var_coefs, '[', 'Pr(>|t|)'  )
  ][, l95      := estimate-(1.96*se),
  ][, u95      := estimate+(1.96*se),

  # Cleanup
  ][, covar_sets := factor(covar_sets, levels=..covar_sets, labels=names(..covar_sets))
  ][, var_coefs := NULL
  ][, lm_coefs := NULL
  ][, fmlas := NULL
]
```
```{r lm-plot1, cache=2, cache.depends=digest::digest(data), fig.width=16, fig.height=10}
ggplot(runs[grepl('mPC',ys)], aes(y=estimate, x=covar_sets, color=covar_sets)) +
  geom_point   (position=position_dodge(width=0.2)) +
  geom_errorbar(position=position_dodge(width=0.2), aes(ymin=l95,ymax=u95), width=0) +
  geom_hline(yintercept=0, color="gray") +
  scale_x_discrete(labels=NULL, breaks=NULL) + labs(x=NULL) +
  facet_grid(rows=vars(xs), cols=vars(ys), scales="free")
```
```{r lm-plot2, cache=2, cache.depends=digest::digest(data), fig.width=18, fig.height=3}
ggplot(runs[grepl('hdl',ys)], aes(y=estimate, x=covar_sets, color=covar_sets)) +
  geom_point   (position=position_dodge(width=0.2)) +
  geom_errorbar(position=position_dodge(width=0.2), aes(ymin=l95,ymax=u95), width=0) +
  geom_hline(yintercept=0, color="gray") +
  scale_x_discrete(labels=NULL, breaks=NULL) + labs(x=NULL) +
  facet_grid(cols=vars(xs), scales="free")
```
```{r lm-plot3, cache=2, cache.depends=digest::digest(data), fig.width=12, fig.height=6}
runs[ys=='hdl_log' &
     xs != 'rs295849' &
     covar_sets=='+gPCs'
   ][order(-t)
   ][, xs := factor(xs,levels=xs) # So ggplot doesn't alphabetically order things
] |>
  ggplot(aes(x=xs, y=t)) +
  geom_bar(stat='identity')
```

# Histograms
```{r hist-prep, warning=F}
vars2hist_quant <- c('pa','mod_pa','mvpa','vig_pa', 'age','bmi','ahei_score','dash_score','ses_score','drinks_per_week', 'hdl','hdl_log')
data2hist_quant <- melt(data[, c('exam','sex', ..vars2hist_quant)],
  measure.vars  = vars2hist_quant,
  variable.name = 'var',
  value.name    = 'val'
)[, exam_means := mean(val,na.rm=T),            by=.(var,exam)
][, `exam_-1sd`:= exam_means - sd(val,na.rm=T), by=.(var,exam)
][, `exam_+1sd`:= exam_means + sd(val,na.rm=T), by=.(var,exam)
][, sex_means := mean(val,na.rm=T),           by=.(var,sex)
][, `sex_-1sd`:= sex_means - sd(val,na.rm=T), by=.(var,sex)
][, `sex_+1sd`:= sex_means + sd(val,na.rm=T), by=.(var,sex)
]

vars2hist_categ <- c('race','smoking','income','rs295849')
data2hist_categ <- melt(data[, c('exam','sex', ..vars2hist_categ)],
  measure.vars  = vars2hist_categ,
  variable.name = 'var',
  value.name    = 'val'
)[, val := factor(val,levels=c( # So ggplot doesn't alphabetically order thigns
    '0','1','2',
    'NEVER', 'FORMER', 'CURRENT',
    'BLACK, AFRICAN-AMERICAN', 'CHINESE AMERICAN', 'HISPANIC', 'WHITE, CAUCASIAN',
    'Missing', '< 5000', '5000 - 7999', '8000 - 11999', '12000 - 15999', '16000 - 19999', '20000 - 24999', '25000 - 29999', '30000 - 34999', '35000 - 39999', '40000 - 49999', '50000 - 74999', '75000 - 99999', '100000 +', '100000 - 124999', '125000 - 149999', '150000 OR MORE'
))]
```
```{r hist-quant, cache=T, cache.extra=digest::digest(data2hist_quant), message=F, warning=F, fig.width=16, fig.height=8}
ggplot(data2hist_quant, aes(x=val, fill=sex)) +
  geom_density(alpha=0.3) +
  facet_wrap(vars(var), scales = 'free') +
  geom_vline(aes(xintercept= sex_means, color=sex)) +
  geom_vline(aes(xintercept=`sex_-1sd`, color=sex), linetype='dashed') +
  geom_vline(aes(xintercept=`sex_+1sd`, color=sex), linetype='dashed')

ggplot(data2hist_quant, aes(x=val, fill=exam)) +
  geom_density(alpha=0.3) +
  facet_wrap(vars(var), scales = 'free') +
  geom_vline(aes(xintercept= exam_means, color=exam)) +
  geom_vline(aes(xintercept=`exam_-1sd`, color=exam), linetype='dashed') +
  geom_vline(aes(xintercept=`exam_+1sd`, color=exam), linetype='dashed')
```
```{r hist-categ, cache=T, cache.depend=digest::digest(data2hist_categ), message=F, warning=F, fig.width=16}
ggplot(data2hist_categ, aes(x=val)) + 
  geom_histogram(aes(fill=sex), stat='count', position='dodge', alpha=0.5) +
  facet_wrap(vars(var), nrow=1, scales = 'free') +
  theme_bw() + theme(axis.text.x = element_text(angle=30, hjust=0.9))
```







#### Plot of education x income
```{r, include=F,eval=F}
# Education x Income
as.data.frame(with(analysis.df, table(educ_4lvl.lab, income_4lvl.lab))) %>% 
    ggplot(aes(x=educ_4lvl.lab, y=Freq, group=income_4lvl.lab, fill=income_4lvl.lab)) + 
    geom_bar(stat = "identity", position=position_stack()) + 
    scale_fill_manual(values = c(proj_palettes$pretty_dark[1:5])) 
```

### Diet phenotypes
Total energy
```{r, include=F,eval=F}
energy_plt <- plot_continuous("energy_kcal")
energy_plt
```

Macronutrients (g & %kcal)
```{r, include=F,eval=F}
carb_plt <- plot_continuous("nut_carb_g")
carb_pct_plt <- plot_continuous("nut_carb_pct")
fat_plt <- plot_continuous("nut_fat_g")
fat_pct_plt <- plot_continuous("nut_fat_pct")
prot_plt <- plot_continuous("nut_prot_g")
prot_pct_plt <- plot_continuous("nut_prot_pct")
carb_plt + carb_pct_plt 
fat_plt + fat_pct_plt
prot_plt + prot_pct_plt
```

Fat types (g & %kcal)
```{r, include=F,eval=F}
mufa_plt <- plot_continuous("nut_mufa_g")
mufa_pct_plt <- plot_continuous("nut_mufa_pct")
pufa_plt <- plot_continuous("nut_pufa_g")
pufa_pct_plt <- plot_continuous("nut_pufa_pct")
sfa_plt <- plot_continuous("nut_sfa_g")
sfa_pct_plt <- plot_continuous("nut_sfa_pct")

mufa_plt + mufa_pct_plt
pufa_plt + pufa_pct_plt
sfa_plt + sfa_pct_plt
```

### Carbohydrate quality
```{r, include=F,eval=F}
fib_plt <- plot_continuous("nut_fiber_g")
fibsol_plt <- plot_continuous("nut_fiber_sol_g")
fibinsol_plt <- plot_continuous("nut_fiber_insol_g")
fib_plt 
fibsol_plt + fibinsol_plt

# note: soluble + insoluble = total
```

**Carbohydrate-to-fiber ratio
```{r, include=F,eval=F}
carb2fib_plt <- plot_continuous("carb2fib") 
fib2carb_plt <- plot_continuous("fib2carb") 
carb2fib_plt + fib2carb_plt
```

```{r, include=F,eval=F}
## Add ranges, means/sd, medians
c("Carb-to-Fiber: ", round(quantile(analysis.df$carb2fib, na.rm=T, probs=seq(0,1,0.2), include.lowest=T), 1))
c("Fiber-to-Carb: ", round(quantile(analysis.df$fib2carb, na.rm=T, probs=seq(0,1,0.2), include.lowest=T), 3))

# Create boxplot to identify potential outliers
carb2fib_boxplt <- analysis.df %>% select("carb2fib") %>%
    filter(!is.na("carb2fib")) %>%
    ggplot(aes(x=carb2fib)) + 
    geom_boxplot(outlier.color="red") +
    ggplot_theme_standard_continuous

# Create boxplot to identify potential outliers
fib2carb_boxplt <- analysis.df %>% select("fib2carb") %>%
    filter(!is.na("fib2carb")) %>%
    ggplot(aes(x=fib2carb)) + 
    geom_boxplot(outlier.color="red") +
    ggplot_theme_standard_continuous

carb2fib_boxplt
fib2carb_boxplt
```

```{r, include=F,eval=F}
#carb-to-fib
bounds_5SD <- c(mean(analysis.df$carb2fib, na.rm=T)+c(-5,5)*sd(analysis.df$carb2fib, na.rm=T))
bounds_5to95 <- quantile(analysis.df$carb2fib, na.rm=T, probs = c(0.05, 0.25, 0.5, 0.75, 0.95))

analysis.df %>% 
select("carb2fib") %>%
    filter(!is.na("carb2fib")) %>%
    mutate(gt5SD = ifelse(carb2fib < bounds_5SD[1] | carb2fib > bounds_5SD[2],">5 SD", "ok")) %>%
    mutate(out5to95 = ifelse(carb2fib < bounds_5to95[1] | carb2fib > bounds_5to95[5],">5 to 95%", "ok")) %>%
    reframe(Outliers_carb2fib_gt5SD=n_pct(gt5SD),
           Outliers_carb2fib_out5to95=n_pct(out5to95))


#fib-to-carb
bounds_5SD <- c(mean(analysis.df$fib2carb, na.rm=T)+c(-5,5)*sd(analysis.df$fib2carb, na.rm=T))
bounds_5to95 <- quantile(analysis.df$fib2carb, na.rm=T, probs = c(0.05, 0.25, 0.5, 0.75, 0.95))

analysis.df %>% 
select("fib2carb") %>%
    filter(!is.na("fib2carb")) %>%
    mutate(gt5SD = ifelse(fib2carb < bounds_5SD[1] | fib2carb > bounds_5SD[2],">5 SD", "ok")) %>%
    mutate(out5to95 = ifelse(fib2carb < bounds_5to95[1] | fib2carb > bounds_5to95[5],">5 to 95%", "ok")) %>%
    reframe(Outliers_fib2carb_gt5SD=n_pct(gt5SD),
           Outliers_fib2carb_out5to95=n_pct(out5to95))

```

Note on outliers for Carbohydrate-to-fiber ratio: 
* 13 (0.3%) values are outside mean+/- 5 SDs
* 406 (10%) vauues are outside 5 to 95% of the data

### Covariates
Lifestyle factors
```{r, include=F,eval=F}
smk_plt <- plot_categorical("smoke_stat.lab")
smkyrs_plt <- plot_continuous("smoke_packyrs")
alc_plt <- plot_categorical("alch_currdrnk.lab")
pa_plt <- plot_continuous("physact_mvpa")
smk_plt + smkyrs_plt
alc_plt + pa_plt
```

Health status-related covariates
```{r, include=F,eval=F}
med_t2d_plt <- plot_categorical("med_t2d")
med_htn_plt <- plot_categorical("med_htn")
med_lip_plt <- plot_categorical("med_lip")
med_t2d_plt + med_htn_plt + med_lip_plt
```

# Run basic correlations among diet variables
### Carbohydrate & macronutrients
```{r, include=F,eval=F}
options(repr.plot.width=14, repr.plot.height=5)
options(warn=-1)

# Carbohydrates vs. macros
carbfat_plt<-plot_xyscatter("nut_carb_g", "nut_fat_g")
carbprot_plt<-plot_xyscatter("nut_carb_g", "nut_prot_g")
carbengy_plot <- plot_xyscatter("nut_carb_g", "energy_kcal")
carbfat_plt + carbprot_plt + carbengy_plot

# Carbohydrates vs. types of fat
carbmufa_plt<-plot_xyscatter("nut_carb_g", "nut_mufa_g")
carbpufa_plt<-plot_xyscatter("nut_carb_g", "nut_pufa_g")
carbsfa_plot <- plot_xyscatter("nut_carb_g", "nut_sfa_g")
carbmufa_plt + carbpufa_plt + carbsfa_plot
```

### Carbohydrate & fiber sources
```{r, include=F,eval=F}
carb2fib_plt <- plot_continuous("carb2fib")
carbfib_plt<-plot_xyscatter("nut_carb_g", "nut_fiber_g")
carb2fib_plt + carbfib_plt 

carbfibsol_plt<-plot_xyscatter("nut_carb_g", "nut_fiber_sol_g")
carbfibinsol_plt<-plot_xyscatter("nut_carb_g", "nut_fiber_insol_g")
carbfibsol_plt + carbfibinsol_plt + carbfibcer_plt
```

# Covariate descriptions across education levels
### Demographic, behavioral & lifestyle phenotypes
```{r, include=F,eval=F}
analysis.df %>% 
    group_by(educ_4lvl.lab) %>%
    reframe(
        N = n(),
        Age = mean_sd(age, d=1),
        Female = n_pct(female, level=1),
        BMI = mean_sd(bmi, d=1),
        Smoking_Current = n_pct(smoke_stat.lab, level="Current"),
        Smoking_Former_lt1yr = n_pct(smoke_stat.lab, level="Former, <1 yr"),
        Smoking_Former_gt1yr = n_pct(smoke_stat.lab, level="Former, >1 yr"),
        Smoking_Never = n_pct(smoke_stat.lab, level="Never"),
        PA_MVPA = mean_sd(physact_mvpa),
        Alchohol_Drinker = n_pct(alch_currdrnk.lab, level="Drinker"),
        Medication_t2d = n_pct(med_t2d, level=1),
        Medication_htn = n_pct(med_htn, level=1),
        Medication_lipid = n_pct(med_lip, level=1),
        Fasting_Glucose= mean_sd(glucose),
        HbA1c=mean_sd(hba1c),
        Triglyceride=mean_sd(tg),
        LDL=mean_sd(ldl),
        HDL=mean_sd(hdl)
    ) %>% t()
```

### Diet phenotypes
Macronutrients
```{r, include=F,eval=F}
analysis.df %>% 
    group_by(educ_4lvl.lab) %>%
    reframe(
        N = n(),
        Energy_kcal=mean_sd(energy_kcal),
        Carb_g=mean_sd(nut_carb_g),
        Protein_g=mean_sd(nut_prot_g),
        Fat_g=mean_sd(nut_fat_g),
        MUFA_g=mean_sd(nut_mufa_g),
        PUFA_g=mean_sd(nut_pufa_g),
        SFA_g=mean_sd(nut_sfa_g),
        Carb_pct=mean_sd(nut_carb_pct),
        Protein_pct=mean_sd(nut_prot_pct),
        Fat_pct=mean_sd(nut_fat_pct),
        MUFA_pct=mean_sd(nut_mufa_pct),
        PUFA_pct=mean_sd(nut_pufa_pct),
        SFA_pct=mean_sd(nut_sfa_pct)
    ) %>% t()
```

Carbohydrate Quality
```{r, include=F,eval=F}
analysis.df %>% 
    group_by(educ_4lvl.lab) %>%
    reframe(
        N = n(),
        Carb_to_Fiber=mean_sd(carb2fib),
        Fiber_to_Carb=mean_sd(fib2carb, d=3),
        Fiber_g=mean_sd(nut_fiber_g),
        Fiber_soluble_g=mean_sd(nut_fiber_sol_g),
        Fiber_insoluble_g=mean_sd(nut_fiber_insol_g)
    ) %>% t()
```

## Univariate SNP-mPC and exposure-mPC associations
```{r, include=F,eval=F}
snp_mPC_assoc_df <- lapply(snp_info_df$rsID, function(rsID) {
    g_dominant <- analysis_df_lcms[[rsID]] > 0
    form_str <- paste0("g_dominant ~ ", paste0("mPC", 1:20, collapse=" + "))
    glm(as.formula(form_str), data=analysis_df_lcms, family="binomial") %>%
        broom::tidy()
}) %>%
    setNames(snp_info_df$rsID) %>%
    bind_rows(.id="snp")

exp_mPC_assoc_df <- lapply(exposures, function(e) {
    e <- analysis_df_lcms[[e]]
    form_str <- paste0("e ~ ", paste0("mPC", 1:20, collapse=" + "))
    glm(as.formula(form_str), data=analysis_df_lcms, family="binomial") %>%
        broom::tidy()
}) %>%
    setNames(exposures) %>%
    bind_rows(.id="exposure")
```

```{r, include=F,eval=F}
snp_mPC_assoc_df %>%
    filter(term != "(Intercept)") %>%
    mutate(nlp = -log10(p.value),
           term = factor(term, levels=paste0("mPC", 1:20))) %>%
    ggplot(aes(x=term, y=snp, fill=nlp)) +
    geom_tile() +
    scale_fill_continuous(name=expression(-log[10] * "(p)"))

exp_mPC_assoc_df %>%
    filter(term != "(Intercept)") %>%
    mutate(nlp = -log10(p.value),
           term = factor(term, levels=paste0("mPC", 1:20))) %>%
    ggplot(aes(x=term, y=exposure, fill=nlp)) +
    geom_tile() +
    scale_fill_continuous(name=expression(-log[10] * "(p)"))
```
