```{css echo=F}
pre {
  overflow-x: scroll;
}
pre code {
  white-space: pre;
}
```

# Setup
```{r setup, message=F}
library(data.table)
library(ggplot2)
library(gganimate)
library(gtsummary)
library(parallel)

options(datatable.na.strings=c('NA',''))
data <- fread('analysis_df.csv')[metabolomics_exam==1] # Only Exam1 metabolomics individuals
```

# Summary table of relevant variables
```{r summary-tbl, cache=T}
data[
  ][, .(bmi, race, smoking, income, sex, pa, mod_pa, mvpa, vig_pa, age, ses_score, drinks_per_week, ahei_score, dash_score, rs295849)
  ][, has_metabolomics := is.na(data$TOM_Id)
  ][ is.na(sex), sex := 'Unknown'
  ] |>
  tbl_summary(
    by=sex,
    missing='no', # TODO more elegant way to display # of missing?
    statistic = list(
      all_continuous()  ~ c("{mean} ± {sd}"),
      all_categorical() ~ c("{n} ({p}%)"   )
  )) |>
  add_overall() |> add_n()
```

# Heatmap of mPCs vs. covars
Note that not all samples contributed to the mPCs — only samples with measurements for all metabolomics methods CP/CN/HP/AN could be used in PCA.
```{r hm, cache=2}
heatmap_covars <- c(
  paste0('gPC', 1:11),
  'site', 'season', 'month',
  'age', 'bmi', 'sex', 'race',
  'ses_score', 'ahei_score', 'dash_score',
  'income', 'drinks_per_week', 'smoking')

runs <- CJ(heatmap_covars, mPC=paste0('mPC',1:20), sorted=F)[
  ][, lm_res    := mapply(mPC,heatmap_covars, FUN = \(mPC,covar) lm(data[[mPC]] ~ data[[covar]]))
  ][, anova_res := lapply(lm_res, anova)
  ][, p         := sapply(anova_res, \(res) res$`Pr(>F)`[1])
]
p_mtx <- as.matrix(dcast(runs, heatmap_covars~mPC, value.var='p'), rownames=1)
p_mtx <- p_mtx[,paste0('mPC',1:20)]

# TODO also do other way of doing heatmap, see journal day 241112.
# Covars, as above: lm(mPC ~ one_covar)
# SNPs, glm(snp ~ all_mPCs, family='binomial')

Heatmap(-log10(p_mtx),
  cell_fun = \(j,i,x,y,w,h,col) if(p_mtx[i,j]<0.05) grid.text('*',x,y,gp=gpar(col='white')),
  col = colorRamp2(c(0,10), c('black','steelblue')),
  cluster_columns=F, cluster_rows=F,
  heatmap_legend_param = list(title = 'log10(p)\nfrom lm')
)
```

Define covariate sets we'll want to test
```{r covar-sets}
covar_sets <- list(
  site                    =c('site'                                                                                                                           ),
  `+sex`                  =c('site', 'sex'                                                                                                                    ),
  `+age`                  =c('site', 'sex', 'age'                                                                                                             ),
  `+ses+income`           =c('site', 'sex', 'age', 'ses_score', 'income'                                                                                      ),
  `+drink+smoke+ahei+dash`=c('site', 'sex', 'age', 'ses_score', 'income', 'drinks_per_week', 'smoking', 'ahei_score', 'dash_score'                            ),
  `+race`                 =c('site', 'sex', 'age', 'ses_score', 'income', 'drinks_per_week', 'smoking', 'ahei_score', 'dash_score', 'race'                    ),
  `+gPCs`                 =c('site', 'sex', 'age', 'ses_score', 'income', 'drinks_per_week', 'smoking', 'ahei_score', 'dash_score', 'race', paste0('gPC', 1:5)),
  `-race`                 =c('site', 'sex', 'age', 'ses_score', 'income', 'drinks_per_week', 'smoking', 'ahei_score', 'dash_score',         paste0('gPC', 1:5)))
```

# Inspect Variable Transformations
HDL vs various transformations of mvpa as in `gxe-omics` workspace's `preliminary-analysis.ipynb`
```{r mvpa-tfs, cache=T}
data <- data[
  ][, mvpa_resid       := resid(lm(mvpa       ~ ., data[, .SD, .SDcols=covar_sets$`+gPCs`], na.action=na.exclude)),
  ][, mvpa_log_resid   := resid(lm(mvpa_log   ~ ., data[, .SD, .SDcols=covar_sets$`+gPCs`], na.action=na.exclude)),
  ][, mvpa_wins_resid  := resid(lm(mvpa_wins  ~ ., data[, .SD, .SDcols=covar_sets$`+gPCs`], na.action=na.exclude)),
  ][, mvpa_trunc_resid := resid(lm(mvpa_trunc ~ ., data[, .SD, .SDcols=covar_sets$`+gPCs`], na.action=na.exclude))
]

rbind(data[, .(hdl, mvpa,           f='identity'   ) ],
      data[, .(hdl, mvpa_resid,       'resid'      ) ],
      data[, .(hdl, mvpa_log,         'log'        ) ],
      data[, .(hdl, mvpa_log_resid,   'resid_log'  ) ],
      data[, .(hdl, mvpa_wins,        'winsor'     ) ],
      data[, .(hdl, mvpa_wins_resid,  'resid_wins' ) ],
      data[, .(hdl, mvpa_trunc,       'trunc'      ) ],
      data[, .(hdl, mvpa_trunc_resid, 'resid_trunc') ],
      use.names=F)[
    ][, f := factor(f,levels=unique(f)) ] |>
  ggplot(aes(x=mvpa, y=hdl)) +
  geom_smooth(method = 'gam', formula = y ~ s(x, bs='cs'), se=T, na.rm=T) +
  facet_wrap(vars(f), ncol=2, scales='free')
```

# Run models
Strategy: run all lm models upfront and examine the results afterward.\
One set of models we want to test is `mPC ~ pa_or_snp_variable + covar_set` for each possible combination of mPC1-10, untransformed PA or SNP variable, and covariate set. The point of this set of models is to see which covariates make the largest difference in the relationship between our variables of interest and the metabolomics, to help figure out which covariates to adjust for.\
Another set of models is `hdl_log ~ pa_or_snp_variable_transformed + covar_set`, for each possible combination of PA or SNP variable, and covariate set. The point of this is to see the main effects, and also see which covariates make the largest difference in these main effects.
```{r covar-sets, cache=T, fig.width=16, fig.height=10}
# mPC ~ pa_or_snp + covars
ys <- paste0('mPC',1:9)
xs <- c('rs295849','pa_bin','pa','mod_pa','mvpa','vig_pa')
lm_runs_w_mPC_outcome <- CJ(ys, xs, covar_sets, sorted=F)

# hdl_log ~ transformed_pa_or_snp + covars
ys <- 'hdl_log'
xs <- c('rs295849', 'pa',       'mod_pa',       'mvpa',       'vig_pa',
                    'pa_log',   'mod_pa_log',   'mvpa_log',   'vig_pa_log',
                    'pa_wins',  'mod_pa_wins',  'mvpa_wins',  'vig_pa_wins',
                    'pa_trunc', 'mod_pa_trunc', 'mvpa_trunc', 'vig_pa_trunc')
lm_runs_w_hdl_outcome <- CJ(ys, xs, covar_sets, sorted=F)

runs <- rbind(lm_runs_w_mPC_outcome,
              lm_runs_w_hdl_outcome)
runs <- runs[, fmlas := paste0('scale(',ys,')','~','scale(',xs,')','+',sapply(covar_sets,paste,collapse='+')) |> sapply(formula) ]

runs[
  ][, lm_coefs := mcmapply(fmlas, FUN = \(fmla)
        lm(fmla,data) |> (\(model) summary(model)$coefficients)(),
      SIMPLIFY=F,     # ↑ Pipe into a function to extract only the coefs. Don't need
      mc.cores=4)     #   all of lm's output. Keeping everything eats up a lot of memory.
  # Extract the results we want
  ][, var_coefs := mapply(lm_coefs, xs, FUN = \(coefs,x)
        coefs[grepl(x,rownames(coefs)),],
      SIMPLIFY=F)

  ][, estimate := sapply(var_coefs, '[', 'Estimate'  )
  ][, se       := sapply(var_coefs, '[', 'Std. Error')
  ][, t        := sapply(var_coefs, '[', 't value'   )
  ][, p        := sapply(var_coefs, '[', 'Pr(>|t|)'  )
  ][, l95      := estimate-(1.96*se),
  ][, u95      := estimate+(1.96*se),

  # Cleanup
  ][, covar_sets := factor(covar_sets, levels=..covar_sets, labels=names(..covar_sets))
  ][, var_coefs := NULL
  ][, lm_coefs := NULL
  ][, fmlas := NULL
]

ggplot(runs[grepl('mPC',ys)], aes(y=estimate, x=covar_sets, color=covar_sets)) +
  geom_point   (position=position_dodge(width=0.2)) +
  geom_errorbar(position=position_dodge(width=0.2), aes(ymin=l95,ymax=u95), width=0) +
  geom_hline(yintercept=0, color="gray") +
  scale_x_discrete(labels=NULL, breaks=NULL) + labs(x=NULL) +
  facet_grid(rows=vars(xs), cols=vars(ys), scales="free")

ggplot(runs[grepl('hdl',ys)], aes(y=estimate, x=covar_sets, color=covar_sets)) +
  geom_point   (position=position_dodge(width=0.2)) +
  geom_errorbar(position=position_dodge(width=0.2), aes(ymin=l95,ymax=u95), width=0) +
  geom_hline(yintercept=0, color="gray") +
  scale_x_discrete(labels=NULL, breaks=NULL) + labs(x=NULL) +
  facet_grid(cols=vars(xs), scales="free")

runs[ys=='hdl_log' &
     xs != 'rs295849' &
     covar_sets=='+race'
   ][order(-t)
   ][, xs := factor(xs,levels=xs) # So that ggplot doesn't alphabetically order it
] |>
  ggplot(aes(x=xs, y=t)) +
  geom_bar(stat='identity')
```

# Histograms
```{r hist-prep, warning=F}
vars2hist_quant <- c('bmi','pa','mod_pa','mvpa','vig_pa','age','ses_score','drinks_per_week','ahei_score','dash_score','hdl')
data4hists_quant <- melt(data[, c('sex', ..vars2hist_quant)],
  measure.vars  = vars2hist_quant,
  variable.name = 'var',
  value.name    = 'val'
)[, means := mean(val,na.rm=T),       by=.(var,sex)
][, `-1sd`:= means - sd(val,na.rm=T), by=.(var,sex)
][, `+1sd`:= means + sd(val,na.rm=T), by=.(var,sex)
]

vars2hist_categ <- c('race','smoking','income_cat','rs295849')
data4hists_categ <- melt(data[, c('sex', ..vars2hist_categ)],
  measure.vars  = vars2hist_categ,
  variable.name = 'var',
  value.name    = 'val'
)
```
```{r hist-quant, cache=2, message=F, warning=F}
ggplot(data4hists_quant, aes(x=val)) + 
  geom_histogram() +
  facet_wrap(vars(var), scales = 'free') +
  geom_vline(aes(xintercept= means), color='red') +
  geom_vline(aes(xintercept=`-1sd`), linetype='dashed') +
  geom_vline(aes(xintercept=`+1sd`), linetype='dashed') +

  transition_states(sex, transition_length = 0.2, state_length = 0.8) +
  ggtitle('Sex: {closest_state}') +
  ease_aes('sine-in-out') +
  theme_bw()
```
```{r hist-categ, cache=2, message=F, warning=F}
ggplot(data4hists_categ, aes(x=val)) + 
  geom_histogram(stat='count') +
  facet_wrap(vars(var), scales = 'free') +

  transition_states(sex, transition_length = 0.2, state_length = 0.8) +
  ggtitle('Sex: {closest_state}') +
  ease_aes('sine-in-out') +
  theme_bw() + theme(axis.text.x = element_text(angle=30, hjust=0.9))
```







```{r}
library(tidyverse)
library(patchwork, table1)
```
```{r, include=F,eval=F}
theme_set(theme_bw())

## ggplot specifications ----------
ggplot_theme_standard_continuous <- theme_bw() + theme(
    axis.text.x = element_text(size=12, vjust=0.65, color = "black"),
    axis.text.y = element_text(size=12, color="black"), 
    strip.text=element_text(size=14, face="bold"),
    axis.title = element_text(size=14, color = "black")
)

ggplot_theme_standard_categorical <- theme_bw() + theme(
    axis.text.x = element_text(size=12, color = "black", angle=30, hjust=0.9),
    axis.text.y = element_text(size=12, color="black"), 
    strip.text=element_text(size=14, face="bold"),
    axis.title = element_text(size=14, color = "black")
)

## Color palettes
proj_palettes <- list(
    greens = paletteer::paletteer_dynamic("cartography::green.pal", 10),
    blues = paletteer::paletteer_dynamic("cartography::blue.pal", 10),
    oranges = paletteer::paletteer_dynamic("cartography::orange.pal", 10),
    pretty_dark = paletteer::paletteer_d("PrettyCols::Dark", 5))
  
```

## Build basic functions
Function to remove outliers by SD
```{r, include=F,eval=F}
remove_outliers.fun <- function(x, SDs=5) {
    bounds <- mean(x, na.rm=T) + SDs * c(-1, 1) * sd(x, na.rm=T)
    
    print(paste0(sum(x < bounds[1], na.rm=TRUE), " outliers removed at <", SDs, " SDs"))
    print(paste0(sum(x > bounds[2], na.rm=TRUE), " outliers removed at >", SDs, " SDs"))
    
    x <- ifelse(x>bounds[1] & x<bounds[2], x, NA) ; x    
}
```

Functions for describing data
```{r, include=F,eval=F}
# Print mean/sd
mean_sd<-function(x, d=2) {
  sprintf("%s \u00B1 %s", round(mean(x, na.rm=T), digits = d), 
          round(sd(x, na.rm=T), digits = d))
}

# Repeat from analysis_prep.Rmd
summarize_continuous <- function(x) {
  m <- round(mean(x, na.rm=TRUE), 1)
  s <- round(sd(x, na.rm=TRUE), 1)
  paste0(m, " (", s, ")")
}

# Print n_pct for categorical vars
n_pct <- function(x, level=F) {
  if(level==F) {
  sapply(as.list(names(table(x))), function(lvl) {
    paste0(lvl, ", ", sum(x == lvl, na.rm=T), " (", round(sum(x == lvl, na.rm=T)/n()*100,1), "%)") }) } 
  else{paste0(sum(x == level, na.rm=T), " (", round(sum(x == level, na.rm=T)/n()*100,1), "%)")}
}

```

```{r, include=F,eval=F}
# Cohort basic descriptives
pop_description_tbl <- analysis.df %>%
    group_by(RaceEthn) %>%
    reframe(
        N = n(),
        Age = mean_sd(age, d=1),
        Female = n_pct(female, level=1),
        BMI = mean_sd(bmi, d=1)) %>%
  arrange(desc(N))
pop_description_tbl %>% t()

# Repeat from analysis_prep.Rmd
pop_description_tbl <- analysis_df %>%
    rename(Race=race) %>%
  group_by(Race) %>%
  summarise(
    N = n(),
    Gender = paste(round(sum(gender_f0m1 == 0) / n() * 100, 1), "% female"),
    Age = summarize_continuous(age),
    BMI = summarize_continuous(bmi),
    `Intentional PA` = paste0(round(sum(pa_bin == 1, na.rm=TRUE) / n() * 100, 1), "%"),
    `Smoking (current)` = paste0(round(sum(smoking == "CURRENT", na.rm=TRUE) / n() * 100, 1), "%"),
    `Smoking (former)` = paste0(round(sum(smoking == "FORMER", na.rm=TRUE) / n() * 100, 1), "%"),
    `PCA-based measure of SES` = summarize_continuous(ses_score),
    
  ) %>%
  arrange(desc(N))
pop_description_tbl %>% t()
```

## Distributions
```{r, include=F,eval=F}
plot_continuous <- function(cont_var) {
  d_complete <- analysis.df %>% select(var=all_of(cont_var)) %>% filter(!is.na(var)) 
  d_complete %>% 
    ggplot(aes(x=var)) + geom_histogram(bins=30) +
    labs(title=cont_var, x=cont_var, y="frequency") +
    geom_vline(xintercept = mean(d_complete$var, na.rm=T), linewidth=1) +
    geom_vline(xintercept = c(mean(d_complete$var)+c(1,-1)*sd(d_complete$var)), linewidth=1, linetype="dashed") + 
    geom_vline(xintercept = median(d_complete$var, na.rm=T), linewidth=2, color = "red") +
    ggplot_theme_standard_continuous
}
plot_continuous <- function(cont_var) { # Repeat from analysis-prep.Rmd
    analysis_df %>%
    filter(!is.na(!!sym(cont_var))) %>%
    ggplot(aes_string(x=cont_var)) +
    geom_histogram(bins=30) +
    labs(title=cont_var)
}

plot_categorical <- function(cat_var) {
  analysis.df %>% 
    select(var=all_of(cat_var)) %>% filter(!is.na(var)) %>%
    ggplot(aes(x=factor(var))) + geom_bar(stat="count") +
    labs(title=cat_var, x=cat_var) +
    ggplot_theme_standard_categorical
}
plot_categorical <- function(cat_var) { # Repeat from analysis-prep.Rmd
    ggplot(analysis_df, aes(x=factor(!!sym(cat_var)))) +
    geom_bar(stat="count") +
    labs(title=cat_var) +
    theme(axis.text.x=element_text(angle=30, hjust=0.9))
}

plot_xyscatter <- function(x_var, y_var) {
  d_complete <- analysis.df %>% select(xvar=x_var, yvar = y_var) %>% filter(complete.cases(.)) 
  d_complete %>%
    ggplot(aes(x=xvar, y=yvar)) + geom_point(size=4, color="#00000075") + 
    labs(title=paste0(x_var, " by ", y_var, "\nr2 = ",
                      round(cor(d_complete$xvar, d_complete$yvar), 3))) +
    ylab(y_var) + xlab(x_var) +
    ggplot_theme_standard_continuous
}

options(repr.plot.width=14, repr.plot.height=5)
options(warn=-1)
```

### Basic descriptives 
```{r, include=F,eval=F}
# Basic descriptives
age_plt <- plot_continuous("age")
gender_plt <- plot_categorical("female")
bmi_plt <- plot_continuous("bmi")
age_plt + gender_plt + bmi_plt

racethn_plt <- plot_categorical("RaceEthn")
racethn_plt

# Repeat from analysis-prep
# Basic biological variables
age_plt <- plot_continuous("age")
gender_plt <- plot_categorical("gender_f0m1")
bmi_plt <- plot_continuous("bmi")
age_plt + gender_plt + bmi_plt

race_plt <- plot_categorical("race")
race_plt
```

### Plots from analysis-prep
```{r, include=F,eval=F}
# Outcomes
hdl_plt <- plot_continuous("hdl")
hdl_log_plt <- plot_continuous("hdl_log")
hdl_plt + hdl_log_plt

hdl_p_plt <- plot_continuous("HDL_P")
m_hdl_p_plt <- plot_continuous("M_HDL_P")
l_hdl_p_plt <- plot_continuous("L_HDL_P")
hdl_p_plt + m_hdl_p_plt + l_hdl_p_plt

# Exposures
pa_plt <- plot_continuous("pa")
pa_bin_plt <- plot_categorical("pa_bin")
pa_plt + pa_bin_plt

mvpa_plt <- plot_continuous("mvpa")
mod_pa_plt <- plot_continuous("mod_pa")
vig_pa_plt <- plot_continuous("vig_pa")
mvpa_plt + mod_pa_plt + vig_pa_plt

analysis_df %>%
    filter(!is.na(vig_pa)) %>%
    ggplot(aes(x=vig_pa, group=gender_f0m1, fill=factor(gender_f0m1))) +
#     geom_density() +
geom_histogram(bins=100) +
    labs(title="vig_pa by gender") +
coord_cartesian(ylim=c(0, 200))

# Batch variables
site_plt <- plot_categorical("site")
month_plt <- plot_categorical("month") + theme(axis.text.x=element_text(angle=30, hjust=0.9))
season_plt <- plot_categorical("season")
site_plt + month_plt + season_plt
```

### SES phenotypes
```{r, include=F,eval=F}
# Education & Income variables
educ_plt <- plot_categorical("educ_lvl.lab")
educ_4lvl_plt <- plot_categorical("educ_4lvl.lab")
educ_plt + educ_4lvl_plt

inc_plt <- plot_categorical("income_lvl.lab")
inc_4lvl_plt <- plot_categorical("income_4lvl.lab")
inc_plt + inc_4lvl_plt

# Somewhat Repeat from analysis-prep.Rmd
# Covariates
ses_plt <- plot_continuous("ses_score")
income_plt <- plot_categorical("income_cat")
ses_plt + income_plt

alc_plt <- plot_continuous("drinks_per_week")
smk_plt <- plot_categorical("smoking")
alc_plt + smk_plt

ahei_plt <- plot_continuous("ahei_score")
dash_plt <- plot_continuous("dash_score")
ahei_plt + dash_plt
```

#### Plot of education x race/ethnicity
```{r, include=F,eval=F}
# make plot for Education levels by R/E
as.data.frame(with(analysis.df, table(RaceEthn, educ_4lvl.lab))) %>% 
    ggplot(aes(x=RaceEthn, y=Freq, group=educ_4lvl.lab, fill=educ_4lvl.lab)) + 
    geom_bar(stat = "identity", position=position_dodge()) + 
    scale_fill_manual(values = c(proj_palettes$pretty_dark[1:5])) + 
    ggplot_theme_standard_categorical

as.data.frame(with(analysis.df, table(RaceEthn, educ_4lvl.lab))) %>% 
    ggplot(aes(x=educ_4lvl.lab, y=Freq, group=RaceEthn, fill=RaceEthn)) + 
    geom_bar(stat = "identity", position=position_dodge()) + 
    scale_fill_manual(values = c(proj_palettes$pretty_dark[1:5])) +
    ggplot_theme_standard_categorical
```

#### Plot of education x income
```{r, include=F,eval=F}
# Education x Income
as.data.frame(with(analysis.df, table(educ_4lvl.lab, income_4lvl.lab))) %>% 
    ggplot(aes(x=educ_4lvl.lab, y=Freq, group=income_4lvl.lab, fill=income_4lvl.lab)) + 
    geom_bar(stat = "identity", position=position_stack()) + 
    scale_fill_manual(values = c(proj_palettes$pretty_dark[1:5])) 
```

### Diet phenotypes
Total energy
```{r, include=F,eval=F}
energy_plt <- plot_continuous("energy_kcal")
energy_plt
```

Macronutrients (g & %kcal)
```{r, include=F,eval=F}
carb_plt <- plot_continuous("nut_carb_g")
carb_pct_plt <- plot_continuous("nut_carb_pct")
fat_plt <- plot_continuous("nut_fat_g")
fat_pct_plt <- plot_continuous("nut_fat_pct")
prot_plt <- plot_continuous("nut_prot_g")
prot_pct_plt <- plot_continuous("nut_prot_pct")
carb_plt + carb_pct_plt 
fat_plt + fat_pct_plt
prot_plt + prot_pct_plt
```

Fat types (g & %kcal)
```{r, include=F,eval=F}
mufa_plt <- plot_continuous("nut_mufa_g")
mufa_pct_plt <- plot_continuous("nut_mufa_pct")
pufa_plt <- plot_continuous("nut_pufa_g")
pufa_pct_plt <- plot_continuous("nut_pufa_pct")
sfa_plt <- plot_continuous("nut_sfa_g")
sfa_pct_plt <- plot_continuous("nut_sfa_pct")

mufa_plt + mufa_pct_plt
pufa_plt + pufa_pct_plt
sfa_plt + sfa_pct_plt
```

### Carbohydrate quality
```{r, include=F,eval=F}
fib_plt <- plot_continuous("nut_fiber_g")
fibsol_plt <- plot_continuous("nut_fiber_sol_g")
fibinsol_plt <- plot_continuous("nut_fiber_insol_g")
fib_plt 
fibsol_plt + fibinsol_plt

# note: soluble + insoluble = total
```

**Carbohydrate-to-fiber ratio
```{r, include=F,eval=F}
carb2fib_plt <- plot_continuous("carb2fib") 
fib2carb_plt <- plot_continuous("fib2carb") 
carb2fib_plt + fib2carb_plt
```

```{r, include=F,eval=F}
## Add ranges, means/sd, medians
c("Carb-to-Fiber: ", round(quantile(analysis.df$carb2fib, na.rm=T, probs=seq(0,1,0.2), include.lowest=T), 1))
c("Fiber-to-Carb: ", round(quantile(analysis.df$fib2carb, na.rm=T, probs=seq(0,1,0.2), include.lowest=T), 3))

# Create boxplot to identify potential outliers
carb2fib_boxplt <- analysis.df %>% select("carb2fib") %>%
    filter(!is.na("carb2fib")) %>%
    ggplot(aes(x=carb2fib)) + 
    geom_boxplot(outlier.color="red") +
    ggplot_theme_standard_continuous

# Create boxplot to identify potential outliers
fib2carb_boxplt <- analysis.df %>% select("fib2carb") %>%
    filter(!is.na("fib2carb")) %>%
    ggplot(aes(x=fib2carb)) + 
    geom_boxplot(outlier.color="red") +
    ggplot_theme_standard_continuous

carb2fib_boxplt
fib2carb_boxplt
```

```{r, include=F,eval=F}
#carb-to-fib
bounds_5SD <- c(mean(analysis.df$carb2fib, na.rm=T)+c(-5,5)*sd(analysis.df$carb2fib, na.rm=T))
bounds_5to95 <- quantile(analysis.df$carb2fib, na.rm=T, probs = c(0.05, 0.25, 0.5, 0.75, 0.95))

analysis.df %>% 
select("carb2fib") %>%
    filter(!is.na("carb2fib")) %>%
    mutate(gt5SD = ifelse(carb2fib < bounds_5SD[1] | carb2fib > bounds_5SD[2],">5 SD", "ok")) %>%
    mutate(out5to95 = ifelse(carb2fib < bounds_5to95[1] | carb2fib > bounds_5to95[5],">5 to 95%", "ok")) %>%
    reframe(Outliers_carb2fib_gt5SD=n_pct(gt5SD),
           Outliers_carb2fib_out5to95=n_pct(out5to95))


#fib-to-carb
bounds_5SD <- c(mean(analysis.df$fib2carb, na.rm=T)+c(-5,5)*sd(analysis.df$fib2carb, na.rm=T))
bounds_5to95 <- quantile(analysis.df$fib2carb, na.rm=T, probs = c(0.05, 0.25, 0.5, 0.75, 0.95))

analysis.df %>% 
select("fib2carb") %>%
    filter(!is.na("fib2carb")) %>%
    mutate(gt5SD = ifelse(fib2carb < bounds_5SD[1] | fib2carb > bounds_5SD[2],">5 SD", "ok")) %>%
    mutate(out5to95 = ifelse(fib2carb < bounds_5to95[1] | fib2carb > bounds_5to95[5],">5 to 95%", "ok")) %>%
    reframe(Outliers_fib2carb_gt5SD=n_pct(gt5SD),
           Outliers_fib2carb_out5to95=n_pct(out5to95))

```

Note on outliers for Carbohydrate-to-fiber ratio: 
* 13 (0.3%) values are outside mean+/- 5 SDs
* 406 (10%) vauues are outside 5 to 95% of the data

### Covariates
Lifestyle factors
```{r, include=F,eval=F}
smk_plt <- plot_categorical("smoke_stat.lab")
smkyrs_plt <- plot_continuous("smoke_packyrs")
alc_plt <- plot_categorical("alch_currdrnk.lab")
pa_plt <- plot_continuous("physact_mvpa")
smk_plt + smkyrs_plt
alc_plt + pa_plt
```

Health status-related covariates
```{r, include=F,eval=F}
med_t2d_plt <- plot_categorical("med_t2d")
med_htn_plt <- plot_categorical("med_htn")
med_lip_plt <- plot_categorical("med_lip")
med_t2d_plt + med_htn_plt + med_lip_plt
```

```{r, include=F,eval=F}
missingness_vars <- c(
    "age", "gender", "bmi", "RaceEthn",
    "educ_4lvl.lab", "income_4lvl.lab",
    "energy_kcal",
    "nut_carb_g", "carb2fib", "nut_fat_g", "nut_prot_g",
    "alch_currdrnk.lab", "smoke_stat.lab",
    "med_t2d", "med_htn", "med_lip")

analysis.df %>%
    select(all_of(missingness_vars)) %>%
    mutate(across(everything(), is.na)) %>%
    summarise(across(everything(), sum)) %>% t()
```

```{r, include=F,eval=F}
# Re-run basic participant descriptives

analysis.df %>%
    group_by(RaceEthn) %>%
    reframe(
        N = n(),
        Age = mean_sd(age, d=1),
        Female = n_pct(female, level=1),
        BMI = mean_sd(bmi, d=1)) %>%
      arrange(desc(N)) %>% t()
```

# Run basic correlations among diet variables
### Carbohydrate & macronutrients
```{r, include=F,eval=F}
options(repr.plot.width=14, repr.plot.height=5)
options(warn=-1)

# Carbohydrates vs. macros
carbfat_plt<-plot_xyscatter("nut_carb_g", "nut_fat_g")
carbprot_plt<-plot_xyscatter("nut_carb_g", "nut_prot_g")
carbengy_plot <- plot_xyscatter("nut_carb_g", "energy_kcal")
carbfat_plt + carbprot_plt + carbengy_plot

# Carbohydrates vs. types of fat
carbmufa_plt<-plot_xyscatter("nut_carb_g", "nut_mufa_g")
carbpufa_plt<-plot_xyscatter("nut_carb_g", "nut_pufa_g")
carbsfa_plot <- plot_xyscatter("nut_carb_g", "nut_sfa_g")
carbmufa_plt + carbpufa_plt + carbsfa_plot
```

### Carbohydrate & fiber sources
```{r, include=F,eval=F}
carb2fib_plt <- plot_continuous("carb2fib")
carbfib_plt<-plot_xyscatter("nut_carb_g", "nut_fiber_g")
carb2fib_plt + carbfib_plt 

carbfibsol_plt<-plot_xyscatter("nut_carb_g", "nut_fiber_sol_g")
carbfibinsol_plt<-plot_xyscatter("nut_carb_g", "nut_fiber_insol_g")
carbfibsol_plt + carbfibinsol_plt + carbfibcer_plt
```

# Covariate descriptions across education levels
### Demographic, behavioral & lifestyle phenotypes
```{r, include=F,eval=F}
analysis.df %>% 
    group_by(educ_4lvl.lab) %>%
    reframe(
        N = n(),
        Age = mean_sd(age, d=1),
        Female = n_pct(female, level=1),
        BMI = mean_sd(bmi, d=1),
        Smoking_Current = n_pct(smoke_stat.lab, level="Current"),
        Smoking_Former_lt1yr = n_pct(smoke_stat.lab, level="Former, <1 yr"),
        Smoking_Former_gt1yr = n_pct(smoke_stat.lab, level="Former, >1 yr"),
        Smoking_Never = n_pct(smoke_stat.lab, level="Never"),
        PA_MVPA = mean_sd(physact_mvpa),
        Alchohol_Drinker = n_pct(alch_currdrnk.lab, level="Drinker"),
        Medication_t2d = n_pct(med_t2d, level=1),
        Medication_htn = n_pct(med_htn, level=1),
        Medication_lipid = n_pct(med_lip, level=1),
        Fasting_Glucose= mean_sd(glucose),
        HbA1c=mean_sd(hba1c),
        Triglyceride=mean_sd(tg),
        LDL=mean_sd(ldl),
        HDL=mean_sd(hdl)
    ) %>% t()
```

### Diet phenotypes
Macronutrients
```{r, include=F,eval=F}
analysis.df %>% 
    group_by(educ_4lvl.lab) %>%
    reframe(
        N = n(),
        Energy_kcal=mean_sd(energy_kcal),
        Carb_g=mean_sd(nut_carb_g),
        Protein_g=mean_sd(nut_prot_g),
        Fat_g=mean_sd(nut_fat_g),
        MUFA_g=mean_sd(nut_mufa_g),
        PUFA_g=mean_sd(nut_pufa_g),
        SFA_g=mean_sd(nut_sfa_g),
        Carb_pct=mean_sd(nut_carb_pct),
        Protein_pct=mean_sd(nut_prot_pct),
        Fat_pct=mean_sd(nut_fat_pct),
        MUFA_pct=mean_sd(nut_mufa_pct),
        PUFA_pct=mean_sd(nut_pufa_pct),
        SFA_pct=mean_sd(nut_sfa_pct)
    ) %>% t()
```

Carbohydrate Quality
```{r, include=F,eval=F}
analysis.df %>% 
    group_by(educ_4lvl.lab) %>%
    reframe(
        N = n(),
        Carb_to_Fiber=mean_sd(carb2fib),
        Fiber_to_Carb=mean_sd(fib2carb, d=3),
        Fiber_g=mean_sd(nut_fiber_g),
        Fiber_soluble_g=mean_sd(nut_fiber_sol_g),
        Fiber_insoluble_g=mean_sd(nut_fiber_insol_g)
    ) %>% t()
```



# BELOW IS WHAT OTHER STUFF THERE WAS LEFT IN ANALYSIS-PREP.ipynb


# Covariate associations with metabolomic PCs
```{r, include=F,eval=F}
race_plt <- bind_cols(analysis_df_lcms, pc_res$x[, 1:5]) %>%
    ggplot(aes(x=PC1, y=PC2)) +
    geom_point(aes(color=race), alpha=0.5)

site_plt <- bind_cols(analysis_df_lcms, pc_res$x[, 1:5]) %>%
    ggplot(aes(x=PC1, y=PC2)) +
    geom_point(aes(color=site), alpha=0.5)

gender_plt <- bind_cols(analysis_df_lcms, pc_res$x[, 1:5]) %>%
    ggplot(aes(x=PC1, y=PC2)) +
    geom_point(aes(color=gender1), alpha=0.5)

options(repr.plot.width=12, repr.plot.height=3)

race_plt + site_plt + gender_plt
```

```{r, include=F,eval=F}
cont_vars <- c("age", "bmi", 
#                paste0("prop_", ancestry_prop_fields), 
               paste0("gPC", 1:11))
bin_vars <- c("smoking", "gender_f0m1")
cat_vars <- c("race", "site", "season", "month")

pc_test <- function(pc_var, covariate) {
    lm_res <- lm(pc_res$x[, pc_var] ~ analysis_df_lcms[[covariate]])
    model_p <- anova(lm_res)[1, 5]
    model_p
}

pc_test_res <- expand_grid(
    pc = paste0("PC", 1:20),
    covar = c(cont_vars, bin_vars, cat_vars)
) %>%
    rowwise() %>%
    mutate(p = pc_test(pc, covar))
```

```{r, include=F,eval=F}
pc_test_plt_df <- pc_test_res %>%
    mutate(nlp = -log10(p),
           sig = ifelse(p < 0.05, "*", ""),
           pc = factor(pc, levels=paste0("PC", 1:20)))

options(repr.plot.width=12, repr.plot.height=6)

pc_test_plt_df %>%
    ggplot(aes(x=pc, y=covar, fill=nlp)) +
    geom_tile() +
    geom_text(aes(label=sig), color="white") +
    scale_fill_continuous(name=expression(-log[10] * "(p)")) +
    scale_size_area(max_size=1.5) +
    labs(title="Covariate-mPC association p-values")

pc_test_plt_df %>%
    mutate(nlp=pmin(nlp, 10)) %>%
    ggplot(aes(x=pc, y=covar, fill=nlp)) +
    geom_tile() +
    geom_text(aes(label=sig), color="white") +
    scale_fill_continuous(name=expression(-log[10] * "(p)")) +
    scale_size_area(max_size=1.5) +
    labs(title="Covariate-mPC association p-values (capped below p<1e-10)")
```

# Assessment of the SNPs and exposures of interest
```{r, include=F,eval=F}
snp_info_df <- read_csv("genotypes/snp_info.csv", col_types=cols())
exposures <- c("pa_bin")
```

## Univariate SNP-mPC and exposure-mPC associations
```{r, include=F,eval=F}
snp_mPC_assoc_df <- lapply(snp_info_df$rsID, function(rsID) {
    g_dominant <- analysis_df_lcms[[rsID]] > 0
    form_str <- paste0("g_dominant ~ ", paste0("mPC", 1:20, collapse=" + "))
    glm(as.formula(form_str), data=analysis_df_lcms, family="binomial") %>%
        broom::tidy()
}) %>%
    setNames(snp_info_df$rsID) %>%
    bind_rows(.id="snp")

exp_mPC_assoc_df <- lapply(exposures, function(e) {
    e <- analysis_df_lcms[[e]]
    form_str <- paste0("e ~ ", paste0("mPC", 1:20, collapse=" + "))
    glm(as.formula(form_str), data=analysis_df_lcms, family="binomial") %>%
        broom::tidy()
}) %>%
    setNames(exposures) %>%
    bind_rows(.id="exposure")
```

```{r, include=F,eval=F}
snp_mPC_assoc_df %>%
    filter(term != "(Intercept)") %>%
    mutate(nlp = -log10(p.value),
           term = factor(term, levels=paste0("mPC", 1:20))) %>%
    ggplot(aes(x=term, y=snp, fill=nlp)) +
    geom_tile() +
    scale_fill_continuous(name=expression(-log[10] * "(p)"))

exp_mPC_assoc_df %>%
    filter(term != "(Intercept)") %>%
    mutate(nlp = -log10(p.value),
           term = factor(term, levels=paste0("mPC", 1:20))) %>%
    ggplot(aes(x=term, y=exposure, fill=nlp)) +
    geom_tile() +
    scale_fill_continuous(name=expression(-log[10] * "(p)"))
```


TODO: try to squish the "Incorporation of technical" AND "biological" covariates sections down into one?
## Incorporation of technical covariates
```{r, include=F,eval=F}
technical_covar_sets <- list(
    none = "",
    add_site = c("gender_f0m1", "age", "site")
)

test_univariate <- function(y, x, covar_vec) {
    form_str <- paste0(y, " ~ ", x)
    if (!identical(covar_vec, "")) {
        form_str <- paste0(form_str, " + ", paste(covar_vec, collapse=" + "))
    }
    lm(as.formula(form_str), data=analysis_df_lcms) %>%
        broom::tidy() %>%
        filter(term == x)
}
```

```{r, include=F,eval=F}
snp_mPC_technical_covar_assoc_df <- expand_grid(
    snp = snp_info_df$rsID,
    mPC = paste0("mPC", 1:3),
    covar_set = names(technical_covar_sets) 
) %>%
    rowwise() %>%
    mutate(lm_res = list(test_univariate(mPC, snp, technical_covar_sets[[covar_set]]))) %>%
    unnest(lm_res)

exp_mPC_technical_covar_assoc_df <- expand_grid(
    e = exposures,
    mPC = paste0("mPC", 1:3),
    covar_set = names(technical_covar_sets) 
) %>%
    rowwise() %>%
    mutate(lm_res = list(test_univariate(mPC, e, technical_covar_sets[[covar_set]]))) %>%
    unnest(lm_res)
```

```{r, include=F,eval=F}
snp_mPC_technical_covar_assoc_df %>%
    mutate(l95 = estimate - 1.96 * std.error,
           u95 = estimate + 1.96 * std.error,
           covar_set = factor(covar_set, levels=names(technical_covar_sets))) %>%
    ggplot(aes(x=snp, y=estimate, color=covar_set)) +
    geom_point(position=position_dodge(width=0.2)) +
    geom_errorbar(aes(ymin=l95, ymax=u95), 
                  position=position_dodge(width=0.2), width=0.1) +
    geom_hline(yintercept=0, color="gray") +
    facet_wrap(~mPC, ncol=1, scales="free")

exp_mPC_technical_covar_assoc_df %>%
    mutate(l95 = estimate - 1.96 * std.error,
           u95 = estimate + 1.96 * std.error,
           covar_set = factor(covar_set, levels=names(technical_covar_sets))) %>%
    ggplot(aes(x=e, y=estimate, color=covar_set)) +
    geom_point(position=position_dodge(width=0.2)) +
    geom_errorbar(aes(ymin=l95, ymax=u95), 
                  position=position_dodge(width=0.2), width=0.1) +
    geom_hline(yintercept=0, color="gray") +
    facet_wrap(~mPC, ncol=1, scales="free")
```

## Incorporation of biological covariates
```{r, include=F,eval=F}
biological_covar_sets <- list(
    basic                   =   "site",
    add_gender_age          = c("site", "gender_f0m1", "age"),
    add_gender_age_race     = c("site", "gender_f0m1", "age", "race"),
    add_gender_age_ses      = c("site", "gender_f0m1", "age", "ses_score", "income_cat"),
    add_gender_age_ses_HL   = c("site", "gender_f0m1", "age", "ses_score", "income_cat", 
                              "drinks_per_week", "smoking", "ahei_score", "dash_score"),
    add_gender_age_gPC      = c("gender_f0m1", "age", paste0("gPC", 1:5)),
    add_gender_age_race_gPC = c("site", "gender_f0m1", "age", "race", paste0("gPC", 1:5))
)
```

```{r, include=F,eval=F}
snp_mPC_biological_covar_assoc_df <- expand_grid(
    snp = snp_info_df$rsID,
    mPC = paste0("mPC", 1:3),
    covar_set = names(biological_covar_sets) 
) %>%
    rowwise() %>%
    mutate(lm_res = list(test_univariate(snp, mPC, biological_covar_sets[[covar_set]]))) %>%
    unnest(lm_res)

exp_mPC_biological_covar_assoc_df <- expand_grid(
    e = exposures,
    mPC = paste0("mPC", 1:3),
    covar_set = names(biological_covar_sets) 
) %>%
    rowwise() %>%
    mutate(lm_res = list(test_univariate(e, mPC, biological_covar_sets[[covar_set]]))) %>%
    unnest(lm_res)
```

```{r, include=F,eval=F}
snp_mPC_biological_covar_assoc_df %>%
    mutate(l95 = estimate - 1.96 * std.error,
           u95 = estimate + 1.96 * std.error,
           covar_set = factor(covar_set, levels=names(biological_covar_sets))) %>%
    ggplot(aes(x=snp, y=estimate, color=covar_set)) +
    geom_point(position=position_dodge(width=0.2)) +
    geom_errorbar(aes(ymin=l95, ymax=u95), 
                  position=position_dodge(width=0.2), width=0.1) +
    geom_hline(yintercept=0, color="gray") +
    facet_wrap(~mPC, ncol=1, scales="free")

exp_mPC_biological_covar_assoc_df %>%
    mutate(l95 = estimate - 1.96 * std.error,
           u95 = estimate + 1.96 * std.error,
           covar_set = factor(covar_set, levels=names(biological_covar_sets))) %>%
    ggplot(aes(x=e, y=estimate, color=covar_set)) +
    geom_point(position=position_dodge(width=0.2)) +
    geom_errorbar(aes(ymin=l95, ymax=u95), 
                  position=position_dodge(width=0.2), width=0.1) +
    geom_hline(yintercept=0, color="gray") +
    facet_wrap(~mPC, ncol=1, scales="free")
```

# Conclusions
* The most important covariates affecting top PCs are study site and race (highly correlated) as well as gender
* It appears that most of the high-level metabolite associations can be captured with about 9 metabolite PCs (for both genotypes and exposures) 
* So, we want to adjust for PEER factors, and we are OK knowing that they are representing expected variables.
# Export final datasets for analysis
```{r, include=F,eval=F}
write_csv(analysis_df, "analysis/analysis_df.csv")  # All individuals
write_csv(analysis_df_lcms, "analysis/analysis_df_lcms.csv")  # Matched to the LC/MS dataset
saveRDS(mesa_metab_mat, "metabolites/lcms_metabolites.rds")
system(paste0("gsutil cp -R phenotypes analysis metabolites $WORKSPACE_BUCKET/"))
```
