# Preliminaries
```{r}
library(tidyverse)
library(patchwork)

if (!require(foreach)) install.packages("foreach")
library(foreach)
if (!require(iterators)) install.packages("iterators")
library(iterators)
if (!require(mediation)) install.packages("mediation")
library(mediation)
```

```{r}
ws_namespace <- Sys.getenv("WORKSPACE_NAMESPACE")
ws_name <- Sys.getenv("WORKSPACE_NAME")
ws_bucket <- Sys.getenv("WORKSPACE_BUCKET")
```

```{r}
theme_set(theme_bw())
```

## Read in phenotypic, genotypic, and metabolomic data
* Phenotype data come from integrating inputs from dbGaP (/PIC-SURE), MESA investigators, and metadata from metabolomic preprocessing.
* Genotype data come from dbGaP (Freeze 9b TOPMed-wide genotype VCF files).
* Metabolomic data come directly from MESA (originally) followed by an extensive QC and preprocessing effort led by Magdalena Sevilla-Gonzalez and Paul Hanson.
```{r}
analysis_df_lcms <- read_csv("analysis/analysis_df_lcms.csv", col_types=cols())
names(analysis_df_lcms)
```

```{r}
snp_info_df <- read_csv("genotypes/snp_info.csv", col_types=cols())
snps <- c("rs295849", "rs2862183")
exposures <- c("mod_vig_pa_bin", "smoking_current", "smoking_ever")
```

```{r}
metabs <- readRDS("metabolites/analysis_metabolites.rds")
```

```{r}
basic_covars <- c("site", "gender_f0m1", "age")
covar_sets <- list(
    no_gPC = c(basic_covars, "ses_score", "income_cat", 
                              "drinks_per_week", "smoking", "ahei_score", "dash_score"),
    all = c(basic_covars, "ses_score", "income_cat", 
                              "drinks_per_week", "smoking", "ahei_score", "dash_score",
                       paste0("gPC", 1:5))
)
covars <- covar_sets$all
```

### What about using a metabolomic E proxy instead?
```{r}
e_proxy_mwas_res <- read_csv("analysis/e_proxy_mwas_res.csv", show_col_types=FALSE)
metabolite_info_df <- read_csv("PH_files/met_info_v12.csv", show_col_types=FALSE) %>%
    dplyr::select(metabolite=Compound_Id_MESA, HMDB_Id, Name, Method)
```

```{r}
sample_n(filter(metabolite_info_df, !is.na(Name)), 3)
```

```{r}
n_eff_metabs <- scan("analysis/n_eff_metabolites.txt", what=double())

pa_mwas_res <- read_csv("analysis/e_proxy_mwas_res.csv", show_col_types=FALSE) %>%
    filter(exposure == "pa") %>%
    arrange(p.value) %>%
    left_join(metabolite_info_df, by="metabolite")

top_pa_mwas_metabolites <- scan("analysis/top_pa_metabolites.txt", what=character())
top_vig_pa_mwas_metabolites <- scan("analysis/top_vig_pa_metabolites.txt", what=character())
# top_pa_mwas_metabolites <- pa_mwas_res$metabolite[1:100]
```

```{r}
top_pa_mwas_metabolites
```

```{r}
test_g_m_interaction <- function(snp,
                                 m_name, 
                                 df=analysis_df_lcms, 
                                 covars=covar_sets$all,
                                 filter_str=NA) {
    df$m <- metabs[, m_name]
    if (!is.na(filter_str)) df <- filter(df, !!rlang::parse_expr(filter_str))
    form_str <- paste0("hdl_log ~ m * ", snp, " + ", paste(covars, collapse=" + "))
    lm(as.formula(form_str), data=df) %>%
    broom::tidy() %>%
    filter(grepl(":", term))
}

m_int_res_df <- expand_grid(
    metabolite = top_pa_mwas_metabolites,
    snp = snps
) %>%
    rowwise() %>%
    mutate(lm_fit = list(test_g_m_interaction(snp, metabolite))) %>%
    unnest(lm_fit) %>%
    arrange(p.value) %>%
    mutate(q = p.adjust(p.value, method="BH")) %>%
    inner_join(metabolite_info_df, by="metabolite")
head(m_int_res_df, 6)

m_int_res_df <- expand_grid(
    metabolite = top_vig_pa_mwas_metabolites,
    snp = snps
) %>%
    rowwise() %>%
    mutate(lm_fit = list(test_g_m_interaction(snp, metabolite))) %>%
    unnest(lm_fit) %>%
    arrange(p.value) %>%
    mutate(q = p.adjust(p.value, method="BH")) %>%
    inner_join(metabolite_info_df, by="metabolite")
head(m_int_res_df, 6)
```

```{r}
0.05 / 8
```

```{r}
m_int_res_df_female <- expand_grid(
    metabolite = top_vig_pa_mwas_metabolites,
    snp = snps
) %>%
    rowwise() %>%
    mutate(lm_fit = list(test_g_m_interaction(snp, metabolite, 
                                              filter_str="gender_f0m1 == 0"))) %>%
    unnest(lm_fit) %>%
    arrange(p.value) %>%
    mutate(q = p.adjust(p.value, method="BH")) %>%
    inner_join(metabolite_info_df, by="metabolite")
head(m_int_res_df_female, 20)
```

```{r}
a <- read_csv("PH_files/met_info_v12.csv", show_col_types=FALSE)
# a %>% filter(Compound_Id_MESA == "QI1084_cp")
# a %>% filter(MZ > 368, MZ < 369, !is.na(Name))

# a %>% filter(Compound_Id_MESA == "QI8416_hp")
# a %>% filter(MZ > 525, MZ < 535, !is.na(Name))


a %>% filter(Compound_Id_MESA == "QI497_cp")
a %>% filter(MZ > 243, MZ < 244, !is.na(Name))

a %>% filter(Compound_Id_MESA == "QI6162_hp")
a %>% filter(MZ > 330, MZ < 335, !is.na(Name))

# a %>% filter(Compound_Id_MESA == "QI8416_hp")
# a %>% filter(MZ > 525, MZ < 535, !is.na(Name))
```

# Test moderated mediation
```{r}
test_moderated_mediation <- function(x, m, y, w, covars) {
    
    # For now, w (the moderating variable) must be binary (0/1)
    med_form_str <- paste0("m ~ x * w + ", paste(covars, collapse=" + "))
    med_fit <- lm(as.formula(med_form_str), data=analysis_df_lcms)
    out_form_str <- paste0("y ~ m * w + x * w + ", paste(covars, collapse=" + "))
    out_fit <- lm(as.formula(out_form_str), data=analysis_df_lcms)
    med_out <- mediate(med_fit, out_fit, 
                       treat="x", mediator="m",
                       robustSE=TRUE, sims=10)  # N_sims doesn't need to be high for this step
    modmed_out <- test.modmed(med_out, list(w=0), list(w=1), sims=100)
    modmed_out
}
```

```{r}
# test_moderated_mediation(analysis_df_lcms$pa, metabs[, 5], analysis_df_lcms$hdl_log, 
#                          analysis_df_lcms$rs2862183, c("age", "gender_f0m1"))
```
